---
title: "genomic_analysis_markdown"
author: "Prediction3 Team"
date: "2025-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This script take the input genotyping variant call files pulled from T3/Wheat as generated in [Genomic_data_import.Rmd](Data/Genomic_data_import.Rmd). This analysis script generates genomic relationship matrices from that genotype data and uses that as an input into the Wishart EM algorithm to develop a combined GRM for all accessions for selected trials generated from [T3_trial_import_and_filtering.Rmd](Data/T3_trial_import_and_filtering.Rmd). Ultimately, a trial-wise genomic similarity matrix is developed to compare trials and select training data for each test set. 

## Load Packages 
```{r, message=FALSE}
library(devtools) #install.packages("devtools")
devtools::install_github("TriticeaeToolbox/BrAPI.R")
library(BrAPI)
library(tidyverse) #install.packages("tidyverse")
library(janitor) #install.packages("janitor")
library(readr) #install.packages("readr")
library(here) #install.packages("here")
library(vcfR) #install.packages("vcfR")
library(stringr) #install.packages("stringr")
devtools::install_github("cran/CovCombR")
library(CovCombR)
library(rrBLUP) #install.packages("rrBLUP")
library(sommer) #install.packages("sommer")
library(AGHmatrix) #install.packages("AGHmatrix")
#library(asreml)
```

## Set BrAPI Server

```{r}
conn <- createBrAPIConnection("wheat.triticeaetoolbox.org", is_breedbase = TRUE)
```

## Load Genomic Data and Trial Data

VCF files for the genotyping projects pulled from T3 that will be included in this analysis were downloaded in the "Genomic_data_import.Rmd" file and are stored in the /vcf_genotyped_data subdirectory. We will read in the VCF files and convert them to dosage matrices.

First we will create paths to locate the stored vcf files

```{r Load paths to vcf data, message=FALSE, warning=FALSE}
genotyped_trials <- readr::read_csv(here::here("data","genotyped_phenotyped_trials.csv"))
test_trials <- readr::read_csv(here::here("data", "test_trial_metadata.csv"))
vcf_projects <- readr::read_csv(here::here("data","selected_geno_projects.csv"))
selected_projects <- vcf_projects$selected_projects
selected_projects <- selected_projects[selected_projects!="12856"]
vcf_paths <-here::here(
  "data",
  "vcf_genotyped_data",
  paste0("raw_project_", selected_projects, ".vcf"))
```

The following function follows paths to vcf files, determines if the files holds data, extracts the gt matrix, parses cells containing strings from GBS datasets, and stores the dosage scores in a dosage matrix named with the genotyping project id. This function can iterate through a vector of paths and stores the resulting matrices in a list.

```{r Function to extract dosage matrices from vcf file paths}
vcf_to_dosage_list <- function(paths) {
  dosage_list <- list()
 
  for (path in paths) {
    result <- tryCatch({
      message("→ Processing ", basename(path))
      vcf <- vcfR::read.vcfR(path)
     
      # skip empty gt
      if (is.null(vcf@gt) || ncol(vcf@gt) < 2) {
        warning("Skipping ", basename(path), ": no genotype data")
        return(NULL)  # <- use return instead of next
      }
     
      gt <- vcf@gt[, -1]
     
      # parse genotype efficiently
      if (any(grepl(":", gt))) {
        gt <- sub(":.*", "", gt)  # vectorized
        dim(gt) <- dim(vcf@gt[, -1])
      }
     
      # normalize phased → unphased
      gt <- gsub("\\|", "/", gt)

      # convert to numeric dosage
      mat <- matrix(NA_integer_, nrow = nrow(gt), ncol = ncol(gt))
      rownames(mat) <- vcf@fix[, "ID"]
      colnames(mat) <- colnames(vcf@gt)[-1]
     
      # vectorized mapping
      mat[gt %in% c("0/0")] <- 0
      mat[gt %in% c("0/1","1/0")] <- 1
      mat[gt %in% c("1/1")] <- 2
     
      mat <- t(mat)
     
      proj_num <- stringr::str_extract(tools::file_path_sans_ext(basename(path)), "\\d+$")
      dosage_list[[paste0("dosage_", proj_num)]] <- mat
     
      message("✓ Finished dosage_", proj_num)
      NULL  # normal exit for tryCatch
     
    }, error = function(e) {
      warning("Skipping ", basename(path), ": ", e$message)
      NULL  # <- return NULL instead of next
    })
  }
 
  return(dosage_list)
}
```

We can run that function to get a list of the dosage matrices for the genotyping projects we are interested in.

```{r Extract dosage matrices, message=FALSE,cache=TRUE}
# Run
dosage_matrices <- vcf_to_dosage_list(vcf_paths)
```

## Constructing a Trial Genomic Similarity Matrix

### Create and filter GRMs for individual genotyping projects

Now we need to convert all the dosage matrices into GRMs for each genotyping project

```{r Construct genomic relationship matrices, message=FALSE}
grm_list <- lapply(dosage_matrices, rrBLUP::A.mat)
names(grm_list) <- paste0("grm_", stringr::str_extract(names(dosage_matrices), "\\d+$"))
```


To allow the code to run faster, we can filter the GRMs to include only 2 types of accessions i) accessions that are a part of our selected trials and/or ii) accessions that appear in multiple GRMs and will be used to calculate relative genomic relationships between non-overlapping accessions.

First we will determine all accessions included in our selected trials
```{r get trial accession names}
accessions <- conn$wizard("accessions", list(trials = c(genotyped_trials$study_name, test_trials$study_name)))
trial_accessions <- unique(accessions$data$names)
```

Next we will determine which accessions have been genotyped in multiple genotyping projects

```{r determine accession overlap, warning=FALSE}
grm_names <- names(grm_list)

## Determine the number of overlapping accessions between GRMs
accession_overlap <- matrix(NA, nrow = length(grm_names), ncol = length(grm_names), dimnames = list(grm_names, grm_names))
accession_overlap <- as.data.frame(accession_overlap)

for (grm1 in grm_names){
  for (grm2 in grm_names){
    accession_overlap[grm1,grm2]=length(intersect(rownames(grm_list$grm1), rownames(grm_list$grm2)))
  }
}

## Determine the accession names of accessions genotyped across multiple GRMs
overlap_accessions <- c()

for (grm1 in grm_names){
  for (grm2 in grm_names){
    if (grm1 == grm2){
      overlap_accessions=overlap_accessions
    }
    else{
    overlap_accessions <- c(overlap_accessions, intersect(rownames(grm_list[[grm1]]), rownames(grm_list[[grm2]])))
    }
  }
}

overlap_accessions <- unique(overlap_accessions)

```

Now we will construct of vector of all accessions we will retain in our GRMs (included in our selected trials or genotyped across multiple GRMs)

```{r determine accessions to retain in GRMs}
trial_or_overlap_accessions <- union(trial_accessions,overlap_accessions)
```

We can filter our raw GRMs to include only accessions from our trial_or_overlap_accessions vector

```{r filter GRMs, message=FALSE}
filtered_grms <- list()

for (grm in grm_names) {
  mat <- grm_list[[grm]]
  
  filt <- mat[rownames(mat) %in% trial_or_overlap_accessions,
              colnames(mat) %in% trial_or_overlap_accessions,
              drop = FALSE]
  
  if (nrow(filt) > 0 && ncol(filt) > 0) {
    filtered_grms[[paste0("filtered_grm_", stringr::str_extract(grm, "\\d+$"))]] <- filt
  } else {
    message("Skipping GRM '", grm, "' because it is empty after filtering")
  }
}
```


### Standardize and Invert GRMs

We need can also standardize the GRMs by making sure the average value on the diagonals is equal to 1

```{r}
std_grm <- function(grm){
  std_grm <- grm / mean(diag(grm))
  return(std_grm)
}
standardized_grms <- lapply(filtered_grms,std_grm)
```

Now we can create G_inv matrices for all of the filtered GRMs

```{r}
ridge <- 1e-6 # Add a ridge to prevent singularity

for (grm_name in names(filtered_grms)) {
  mat <- filtered_grms[[grm_name]]
  filtered_grms[[grm_name]] <- mat + diag(ridge, nrow(mat))
}

inverted_grms <- lapply(filtered_grms,solve)
names(inverted_grms) <- paste0("filtered_grm_inv_", stringr::str_extract(names(filtered_grms), "\\d+$"))
```

### Combining GRMs

We can combine GRMs of different genotype matrices using a set of functions to run a Wishart EM algorithm developed by Deniz Akdemir [https://github.com/denizakdemir/wishartem/blob/main/wishartem/em_covariance_combiner.R](https://github.com/denizakdemir/wishartem/blob/main/wishartem/em_covariance_combiner.R)

```{r}
#' @title EM Covariance Combiner
#' @description Implementation of EM algorithm for combining partial covariance matrices.
#' This function implements the EM algorithm for combining multiple partially observed
#' covariance matrices into a single combined covariance matrix.
#' 
#' @param partial_covs List of partial covariance matrices
#' @param var_indices List of variable indices for each partial covariance matrix
#' @param degrees_freedom Optional numeric vector of degrees of freedom for each matrix (default: 100 for each)
#' @param max_iter Maximum number of iterations for the EM algorithm (default: 100)
#' @param tol Convergence tolerance for the EM algorithm (default: 1e-6)
#' @param track_loglik Whether to track log-likelihood values during iterations (default: TRUE)
#' 
#' @return List containing:
#'   - psi: The combined covariance matrix
#'   - sampling_cov: Sampling covariance matrix for the combined covariance matrix
#'   - loglik_path: Vector of log-likelihood values at each iteration (if track_loglik=TRUE)
#' 
#' @importFrom Matrix solve
#' @importFrom stats cov
#' @export
EMCovarianceCombiner <- function(partial_covs, var_indices, degrees_freedom = NULL,
                                max_iter = 100, tol = 1e-6, track_loglik = TRUE, compute_sampling_cov = FALSE) {
  
  # Validate inputs
  if (length(partial_covs) == 0 || length(var_indices) == 0) {
    stop("Empty input provided")
  }
  if (length(partial_covs) != length(var_indices)) {
    stop("Number of covariance matrices must match number of index sets")
  }
  
  # Determine total number of variables
  n_vars <- max(sapply(var_indices, function(idx) max(idx))) + 1
  
  # Set default degrees of freedom if not provided
  if (is.null(degrees_freedom)) {
    degrees_freedom <- rep(100, length(partial_covs))
  }
  if (length(degrees_freedom) != length(partial_covs)) {
    stop("Length of degrees_freedom must match number of matrices")
  }
  
  # Initialize log-likelihood path
  loglik_path <- if (track_loglik) numeric(0) else NULL
  
  # Initialize Psi (combined covariance matrix)
  psi <- initialize_psi(partial_covs, var_indices, n_vars)
  
  # Compute initial log-likelihood if tracking
  if (track_loglik) {
    loglik <- compute_log_likelihood(psi, partial_covs, var_indices, degrees_freedom)
    loglik_path <- c(loglik_path, loglik)
  }
  
  # EM iterations
  pb <- txtProgressBar(min = 0, max = max_iter, style = 3)
  for (iter in 1:max_iter) {
    setTxtProgressBar(pb, iter)
    psi_old <- psi
    
    # E-step: Compute conditional expectations
    expectations <- vector("list", length(partial_covs))
    for (i in seq_along(partial_covs)) {
      ya <- partial_covs[[i]]
      idx <- var_indices[[i]]
      nu <- degrees_freedom[i]
      missing_idx <- setdiff(0:(n_vars-1), idx)
      
      # Calculate conditional expectation
      exp_y <- compute_conditional_expectation(ya, psi, idx, missing_idx)
      expectations[[i]] <- list(exp_y = exp_y, nu = nu)
    }
    
    # M-step: Update Psi
    psi_sum <- matrix(0, nrow = n_vars, ncol = n_vars)
    total_nu <- sum(degrees_freedom)
    
    for (exp in expectations) {
      psi_sum <- psi_sum + exp$nu * exp$exp_y
    }
    psi <- psi_sum / total_nu
    
    # Compute log-likelihood if tracking
    if (track_loglik) {
      loglik <- compute_log_likelihood(psi, partial_covs, var_indices, degrees_freedom)
      loglik_path <- c(loglik_path, loglik)
    }
    
    # Check convergence
    if (max(abs(psi - psi_old)) < tol) {
      message("\nConverged at iteration ", iter)
      break
    }
  }
  close(pb)
  # Compute sampling covariance if requested
if (compute_sampling_cov) {
  message("Computing sampling covariance (this may take a while)...")
  sampling_cov <- tryCatch({
    compute_sampling_covariance(psi, partial_covs, var_indices, degrees_freedom)
  }, error = function(e) {
    warning("Error computing sampling covariance: ", e$message)
    return(NULL)
  })
} else {
  sampling_cov <- NULL
}
  
  # Return results
  return(list(
    psi = psi,
    sampling_cov = sampling_cov,
    loglik_path = loglik_path
  ))
}

#' Initialize the combined covariance matrix using available variances
#' 
#' @param partial_covs List of partial covariance matrices
#' @param var_indices List of variable indices for each partial covariance matrix
#' @param n_vars Total number of variables
#' @return Initialized covariance matrix
initialize_psi <- function(partial_covs, var_indices, n_vars) {
  psi <- diag(n_vars)
  
  for (i in 0:(n_vars-1)) {
    variances <- numeric(0)
    
    for (j in seq_along(partial_covs)) {
      cov <- partial_covs[[j]]
      idx <- var_indices[[j]]
      
      if (i %in% idx) {
        i_local <- match(i, idx) - 1  # Convert to 0-based index
        variances <- c(variances, cov[i_local + 1, i_local + 1])  # Convert back to 1-based for R
      }
    }
    
    if (length(variances) > 0) {
      psi[i + 1, i + 1] <- mean(variances)  # 1-based indexing in R
    }
  }
  
  return(psi)
}

#' Compute conditional expectation for the E-step of the EM algorithm
#' 
#' @param ya Observed partial covariance matrix
#' @param psi Current estimate of the combined covariance matrix
#' @param obs_idx Indices of observed variables (0-based)
#' @param missing_idx Indices of missing variables (0-based)
#' @return Conditional expectation of the complete covariance matrix
compute_conditional_expectation <- function(ya, psi, obs_idx, missing_idx) {
  # Convert to 1-based indices for R
  obs_idx_r <- obs_idx + 1
  missing_idx_r <- missing_idx + 1
  
  n_total <- nrow(psi)
  result <- matrix(0, nrow = n_total, ncol = n_total)
  
  # If there are no missing indices, just expand ya to full dimension
  if (length(missing_idx) == 0) {
    result[obs_idx_r, obs_idx_r] <- ya
    return(result)
  }
  
  # Extract submatrices from psi and bend to avoid singularity issues in solving
  psi_aa <- psi[obs_idx_r, obs_idx_r, drop = FALSE]
  psi_aa <- psi_aa + diag(1e-5, nrow(psi_aa))
  psi_ab <- psi[obs_idx_r, missing_idx_r, drop = FALSE]
  psi_bb <- psi[missing_idx_r, missing_idx_r, drop = FALSE]
  
  # Compute conditional expectations efficiently
  # Use Matrix::solve for potential speedup with sparse matrices
  psi_aa_inv <- solve(psi_aa)
  b_matrix <- t(psi_ab) %*% psi_aa_inv
  
  # Fill observed part
  result[obs_idx_r, obs_idx_r] <- ya
  
  # Fill cross-terms
  exp_yab <- b_matrix %*% ya
  result[missing_idx_r, obs_idx_r] <- exp_yab
  result[obs_idx_r, missing_idx_r] <- t(exp_yab)
  
  # Fill missing part
  result[missing_idx_r, missing_idx_r] <- psi_bb - b_matrix %*% psi_ab + b_matrix %*% ya %*% t(b_matrix)
  
  return(result)
}

#' Compute the observed data log-likelihood
#' 
#' @param psi Current estimate of the combined covariance matrix
#' @param partial_covs List of partial covariance matrices
#' @param var_indices List of variable indices for each partial covariance matrix
#' @param degrees_freedom Degrees of freedom for each partial covariance matrix
#' @return The observed data log-likelihood
compute_log_likelihood <- function(psi, partial_covs, var_indices, degrees_freedom) {
  log_lik <- 0.0
  
  for (i in seq_along(partial_covs)) {
    ya <- partial_covs[[i]]
    idx <- var_indices[[i]]
    nu <- degrees_freedom[i]
    
    # Convert to 1-based indices for R
    idx_r <- idx + 1
    
    # Extract the submatrix of psi corresponding to the observed variables
    psi_subset <- psi[idx_r, idx_r, drop = FALSE]
    psi_subset <- psi_subset + diag(1e-6, nrow(psi_subset))
    
    # Check if matrix is positive definite
    eig <- try(eigen(psi_subset, symmetric = TRUE, only.values = TRUE)$values, silent = TRUE)
    if (inherits(eig, "try-error") || any(eig <= 0)) {
      return(-Inf)  # Non-positive definite matrix
    }
    
    # Compute log-likelihood contribution (Wishart log-likelihood, ignoring constant terms)
    logdet_psi <- sum(log(eig))  # More efficient than determinant()
    
    # Trace term using matrix inverse
    psi_inv <- solve(psi_subset)
    trace_term <- sum(diag(psi_inv %*% ya))
    
    # Add to log-likelihood (scaled by degrees of freedom)
    log_lik <- log_lik - 0.5 * nu * (logdet_psi + trace_term)
  }
  
  return(log_lik)
}

#' Map matrix indices to vector index for covariance parameters
#' 
#' @param i Row index (0-based)
#' @param j Column index (0-based)
#' @return Vector index
matrix_to_vector_idx <- function(i, j) {
  if (i < j) {
    # Swap to ensure i >= j
    tmp <- i
    i <- j
    j <- tmp
  }
  return(i * (i + 1) / 2 + j)
}

#' Compute the sampling covariance matrix for the combined covariance matrix
#' 
#' @param psi Combined covariance matrix
#' @param partial_covs List of partial covariance matrices
#' @param var_indices List of variable indices for each partial covariance matrix
#' @param degrees_freedom Degrees of freedom for each partial covariance matrix
#' @return Sampling covariance matrix for the combined covariance matrix
compute_sampling_covariance <- function(psi, partial_covs, var_indices, degrees_freedom) {
  n_vars <- nrow(psi)
  n_params <- n_vars * (n_vars + 1) / 2  # Number of unique elements
  
  # Initialize matrices for sandwich formula
  information <- matrix(0, nrow = n_params, ncol = n_params)
  score_cov <- matrix(0, nrow = n_params, ncol = n_params)
  
  # Compute the inverse of psi once
  psi_inv <- solve(psi)
  
  # Convert indices to 0-based for consistency with the matrix_to_vector_idx function
  for (i in seq_along(partial_covs)) {
    ya <- partial_covs[[i]]
    idx <- var_indices[[i]]  # 0-based
    nu <- degrees_freedom[i]
    
    missing_idx <- setdiff(0:(n_vars-1), idx)
    
    # Convert to 1-based for R computation
    idx_r <- idx + 1
    missing_idx_r <- missing_idx + 1
    
    # Get conditional expectation
    exp_y <- compute_conditional_expectation(ya, psi, idx, missing_idx)
    
    # Compute score contribution
    score <- exp_y - psi
    
    # Flatten the matrices efficiently
    score_vec <- numeric(n_params)
    
    # Map the score matrix to vector form
    for (i_idx in 0:(n_vars-1)) {
      for (j_idx in 0:i_idx) {
        idx_vec <- matrix_to_vector_idx(i_idx, j_idx) + 1  # +1 for R indexing
        score_vec[idx_vec] <- score[i_idx + 1, j_idx + 1]  # +1 for R indexing
      }
    }
    
    # Update score covariance scaled by nu
    score_cov <- score_cov + nu * tcrossprod(score_vec)
    
    # Compute the observed information matrix contribution - use efficient vectorization
    for (i_idx in 0:(n_vars-1)) {
      for (j_idx in 0:i_idx) {
        idx1 <- matrix_to_vector_idx(i_idx, j_idx) + 1  # +1 for R indexing
        
        for (k_idx in 0:(n_vars-1)) {
          for (l_idx in 0:k_idx) {
            idx2 <- matrix_to_vector_idx(k_idx, l_idx) + 1  # +1 for R indexing
            
            # +1 for R indexing
            info_val <- nu * (
              psi_inv[i_idx + 1, k_idx + 1] * psi_inv[j_idx + 1, l_idx + 1] +
              psi_inv[i_idx + 1, l_idx + 1] * psi_inv[j_idx + 1, k_idx + 1]
            )
            
            information[idx1, idx2] <- information[idx1, idx2] + info_val
          }
        }
      }
    }
  }
  
  # Compute sandwich covariance matrix
  info_inv <- solve(information)
  sampling_cov <- info_inv %*% score_cov %*% t(info_inv)
  
  return(sampling_cov)
}

```

To run the algorithm, we need a numerical index value for each germplasm entry in all of the GRMs. We will just pull the T3 germplasm ID values.

```{r}
accession_index_map <- data.frame(
  accession_name = unique(unlist(lapply(filtered_grms, colnames))),
  accession_index = seq(0, length(unique(unlist(lapply(filtered_grms, colnames)))) - 1),
  stringsAsFactors = FALSE
)

# Initialize lists to store names and indices
grm_names_list <- list()
grm_indices_list <- list()

for (grm_name in names(filtered_grms)) {
  mat <- filtered_grms[[grm_name]]
  colnames_mat <- colnames(mat)
  
  # Store accession names
  grm_names_list[[grm_name]] <- colnames_mat
  
  # Store indices in accession_index_map
  grm_indices_list[[grm_name]] <- match(colnames_mat, accession_index_map$accession_name)
}
```

Run the algorithm, extract the combined GRM, and relabel the accessions

```{r}
original_names <- sub("^filtered_", "", names(filtered_grms))

valid_indices <- which(names(grm_list) %in% original_names)

result <- EMCovarianceCombiner(partial_covs = standardized_grms,
                               var_indices = grm_indices_list,
                               degrees_freedom = sapply(dosage_matrices[valid_indices], nrow),
                               compute_sampling_cov = FALSE)
saveRDS(result, here::here("analysis","new_EMcombinedGRM_wDF.rds"))
comb_grm <- result$psi[-1,-1]
rownames(comb_grm) <- accession_index_map$accession_name
colnames(comb_grm) <- accession_index_map$accession_name
saveRDS(comb_grm, here::here("analysis","comb_grm_mat.rds"))
```

Now we want to create a loop that will loop through the trials we are interested in to create a trial-wise genomic similarity matrix (G).

```{r}
trial_names <- c(genotyped_trials$study_name, test_trials$study_name)

accessions_by_trial <- lapply(trial_names, function(trial) {
  conn$wizard("accessions", list(trials = trial))
})

names(accessions_by_trial) <- trial_names
accession_names_by_trial <- lapply(accessions_by_trial, function(x) x$data$names)
accession_names_by_trial <- lapply(accession_names_by_trial, unique)



G <- matrix(data=NA, nrow=length(accession_names_by_trial), ncol=length(accession_names_by_trial), dimnames = list(names(accession_names_by_trial), names(accession_names_by_trial)))

for (trial1 in names(accession_names_by_trial)){
  for(trial2 in names(accession_names_by_trial)){
    
    ## pull accessions for each trial
    trial1_acc <-accession_names_by_trial[[trial1]]
    trial2_acc <- accession_names_by_trial[[trial2]]
    
    # Keep only accessions found in comb_grm
    common_acc1 <- intersect(accession_names_by_trial[[trial1]], rownames(comb_grm))
    common_acc2 <- intersect(accession_names_by_trial[[trial2]], colnames(comb_grm))

    # Take only accessions present in both sets
    pairwise_rel <- comb_grm[common_acc1, common_acc2, drop = FALSE]
    average_rel <- mean(pairwise_rel, na.rm = TRUE)
    G[trial1,trial2] <- average_rel
  }
}
keep <- !is.na(diag(G))

G_clean <- G[keep, keep]

rescaled_G <- G_clean / mean(diag(G_clean))

saveRDS(rescaled_G, "trial_Gmat.rds")



```

Plot G as a heatmap

```{r}
wsu_trials <- genotyped_trials |> dplyr::filter(program_name=="Washington State University")
wsu_studies <- wsu_trials$study_name

wsu_Gmat <- rescaled_G[wsu_studies, wsu_studies]

non_wsu_Gmat <- rescaled_G[!rownames(rescaled_G) %in% wsu_studies,!colnames(rescaled_G) %in%wsu_studies]

# Compute clustering orders (rows and columns)
row_ord <- hclust(dist(non_wsu_Gmat))$order
col_ord <- hclust(dist(t(non_wsu_Gmat)))$order

# Reorder matrix by those cluster orders
mat_ord <- non_wsu_Gmat[row_ord, col_ord, drop = FALSE]

# Plot heatmap with no dendrograms, but ordered by clustering
heatmap(
  mat_ord,
  Rowv = NA,  # no dendrogram
  Colv = NA,  # no dendrogram
  scale = "none",
  cexRow = 1.2,
  cexCol = 1,
  margins = c(15, 15),
  main = "Average Genomic Similarity of T3 Trials (G_rel)",
)
  

row_ord_wsu <- hclust(dist(wsu_Gmat))$order
col_ord_wsu <- hclust(dist(t(wsu_Gmat)))$order

# Reorder matrix by those cluster orders
mat_ord_wsu <- wsu_Gmat[row_ord_wsu, col_ord_wsu, drop = FALSE]

# Plot heatmap with no dendrograms, but ordered by clustering
heatmap(
  mat_ord_wsu,
  Rowv = NA,  # no dendrogram
  Colv = NA,  # no dendrogram
  scale = "none",
  cexRow = 1.2,
  cexCol = 1,
  margins = c(15, 15),
  main = "Average Genomic Similarity of WSU Trials (G_rel)",
)
```

```{r}
trial_gen_pca <- prcomp(non_wsu_Gmat, scale. = TRUE)

cum_var <- cumsum(trial_gen_pca$sdev^2) / sum(trial_gen_pca$sdev^2)
k <- which(cum_var >= 0.95)[1] #number of PCs to explain variance
trial_gen_pca_scores <- trial_gen_pca$x[,1:k]
pc_centeroid <- colMeans(trial_gen_pca_scores)
trial_gen_sim_dist <- mahalanobis(trial_gen_pca_scores, center = pc_centeroid, cov= cov(trial_gen_pca_scores)) # calculates distance from the centroid of the PCA 
trial_gen_dist <- data.frame(study_db_id = colnames(non_wsu_Gmat), distance = trial_gen_sim_dist)
write_csv(trial_gen_dist, here::here("analysis","trial_gen_distances.csv"))

trial_gen_pca_scores <- as.data.frame(trial_gen_pca_scores)
trial_gen_pca_scores$trial <- rownames(non_wsu_Gmat)
trial_gen_pca_scores <- trial_gen_pca_scores |> 
  left_join(genotyped_trials |> dplyr::select(study_name,program_name), by = c("trial" = "study_name"))
ggplot(trial_gen_pca_scores, aes(PC1,PC2,label=trial, color=program_name))+
  geom_point()+
  geom_text(vjust=-0.5)+
  theme_minimal()+
  labs(title="Genomic PCA of selected non-WSU T3 trials")

# Inspect variance explained
summary(trial_gen_pca)

# Make a biplot
biplot(trial_gen_pca, main = "PCA of Covariance Matrix")
```



