---
title: "GBLUP"
author: "Prediction3 Team"
date: "2026-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This script uses the trial G ([trial_Gmat.rds](Analysis/trial_Gmat.rds)) and W ([Trial_W_mat.csv](Analysis/Trial_W_mat.csv)) matrices to select optimal training data for each test set trial. Functions are developed to run GBLUP and multikernel BLUP models for each test set. The multikernel BLUP model fits random effects for genotype and environment using the combined GRM (output from [genomic_analysis.Rmd](Analysis/genomic_analysis.Rmd)) and trial-wise W matrix as covariance structures. Leave-One-Trial-Out Cross Validation is performed to evaluate prediction accuracies among one-trial subsets of the training data. Functions are developed to generate CV0 and CV00 predictions for the different models. Ultimately,the multikernel BLUP model was used for T3 Predictathon competition submissions.

## Load Packages

```{r load packages, message=FALSE}
library(devtools) #install.packages("devtools")
devtools::install_github("TriticeaeToolbox/BrAPI.R")
library(BrAPI)
library(tidyverse) #install.packages("tidyverse")
library(janitor) #install.packages("janitor")
library(readr) #install.packages("readr")
library(here) #install.packages("here")
library(vcfR) #install.packages("vcfR")
library(stringr) #install.packages("stringr")
devtools::install_github("cran/CovCombR")
library(CovCombR)
library(rrBLUP) #install.packages("rrBLUP")
library(sommer) #install.packages("sommer")
library(AGHmatrix) #install.packages("AGHmatrix")
#library(asreml)
```

## Set BrAPI Connection

```{r set BrAPI connection, message=FALSE}
conn <- createBrAPIConnection("wheat.triticeaetoolbox.org", is_breedbase = TRUE)
```

## Load Data
```{r load data, message=FALSE}
test_trials <- readr::read_csv(here::here("data","test_trial_metadata.csv"))
test_set <- test_trials$study_name
training_trials <- readr::read_csv(here::here("data","genotyped_phenotyped_trials.csv"))
phenotypes <- readr::read_csv(here::here("data","genotyped_trial_phenotypes.csv"))

trial_W <- readr::read_csv(here::here("analysis","trial_W_mat.csv"))
rownames(trial_W) <- colnames(trial_W)
trial_G <- as.data.frame(readr::read_rds(here::here("analysis","trial_Gmat.rds")))

comb_grm <- read_rds(here::here("analysis", "comb_grm_mat.rds"))
```

## Select Training Set Trials for each Test Set

```{r Construct Training Datasets function}
select_training_data <- function(test_trials, test_trial_metadata_df, training_trial_metadata_df, trial_W, trial_G, phenotypes, comb_grm, test_trial_metadata, top_n = 15){
  ## ARGS
  
  ## Returns
  result <- vector("list", length(test_trials))
  names(result) <- test_trials
  
  for (test_trial in test_trials){
    
    test_accessions <- conn$wizard("accessions",filters = list(trials = test_trial))$data$names
    result[[test_trial]]$test_accessions <- test_accessions
    
    result[[test_trial]]$test_trial_metadata <- test_trial_metadata_df |> filter(study_name == test_trial)
    
    trial_W <- as.matrix(trial_W)
    env_similarities <- trial_W[,test_trial]
    env_similarities <- env_similarities[names(env_similarities) != test_trial]
    top_envs <- names(sort(env_similarities, decreasing = TRUE))[1:top_n]
    
    trial_G <- as.matrix(trial_G)
    gen_similarities <- trial_G[,test_trial]
    gen_similarities <- gen_similarities[names(gen_similarities) != test_trial]
    top_gens <- names(sort(gen_similarities, decreasing = TRUE))[1:top_n]
    training_set_trials <-  unique(c(top_envs, top_gens))
    
    result[[test_trial]]$training_set_trials <- training_set_trials
    
    result[[test_trial]]$training_set_metadata <- training_trial_metadata_df |> filter(study_name %in% training_set_trials)
    
    training_data <- phenotypes |> dplyr::filter(study_name %in% training_set_trials)
    result[[test_trial]]$training_data <- training_data
    
    training_accessions <- unique(training_data$germplasm_name)
    result[[test_trial]]$training_accessions <- training_accessions
    
    genotyped_accessions <- intersect(unique(c(test_accessions,training_accessions)), rownames(comb_grm))
    
    subset_grm <- comb_grm[genotyped_accessions,genotyped_accessions]
    result[[test_trial]]$subset_grm <- subset_grm
    
    subset_W <- trial_W[c(test_trial,training_set_trials), c(test_trial,training_set_trials)]
    result[[test_trial]]$subset_W <- subset_W
    
    subset_pheno <- training_data |> 
      dplyr::filter(germplasm_name %in% genotyped_accessions) |> 
      dplyr::select(study_db_id,study_name,germplasm_name,grain_yield_kg_ha_co_321_0001218)
    result[[test_trial]]$subset_pheno <- subset_pheno
    
    test_pheno <- data.frame(
      germplasm_name = test_accessions, 
      study_name = test_trial, 
      study_db_id = dplyr::filter(test_trial_metadata,study_name == test_trial)$study_db_id, 
      grain_yield_kg_ha_co_321_0001218 = NA
    ) |> 
      dplyr::filter(germplasm_name %in% genotyped_accessions)
    result[[test_trial]]$test_pheno <- test_pheno
    
    result[[test_trial]]$combined_pheno <- rbind(subset_pheno,test_pheno)
  }
  
  return(result)
}
```

```{r pull training data}
training_data <- select_training_data(test_trials = test_set, test_trial_metadata_df=test_trials, training_trial_metadata_df = training_trials, trial_W = trial_W, trial_G = trial_G, phenotypes = phenotypes, comb_grm = comb_grm, test_trial_metadata = test_trials)
```

```{r fit GBLUP models}
## Without trial fixed effect
gblup_base <- function(training_data){
  # -----------------------------------
  # DESCRIPTION:
  # Performs genomic prediction (y = mu + ug + e) for a given test set of trials using a provided training data from selected trials
  #
  # ARGS:
  # training_data: list
  #   A list of training data set where each element contains    
  #
  # DEPENDENDCIES:
  #
  #
  # RETURNS: 
  #
  #
  # -----------------------------------
  result <- vector("list", length(training_data))
  names(result) <- names(training_data)
  
  pb <- txtProgressBar(min = 0, max = length(training_data), style = 3)
  
  for (i in seq_along(training_data)){
    test_trial <- names(training_data)[i]
    current_data <- training_data[[test_trial]]
    result[[test_trial]] <- rrBLUP::kin.blup(
      data = as.data.frame(current_data$combined_pheno),
      geno = "germplasm_name",
      pheno = "grain_yield_kg_ha_co_321_0001218",
      K = as.matrix(current_data$subset_grm)
    )
    setTxtProgressBar(pb, i)
  }
  close(pb)
  return(result)
}

```

```{r run base GBLUP}
base_GBLUP <- gblup_base(training_data)
```

```{r GBLUP LOO_CV}
GBLUP_LOO_CV <- function(training_data)
```


Model: $$ y \sim \mu + Z_Gu_G + Z_Eu_E + e$$ 


```{r multikernel GBLUP function}
multi_GBLUP <- function(training_data){
  
  predictions <- vector("list", length(training_data))
  names(predictions) <- names(training_data)
  
  for (test_trial in seq_along(training_data)){
    # Pull data for current test trial from training data list
    current_data <- training_data[[test_trial]]
    y <- current_data$combined_pheno
    test_env <- current_data$test_trial_metadata$study_name
    
    # Get genotypes and environment information for all trials and set as factors
    gens <- unique(y$germplasm_name)
    envs <- unique(y$study_name)
    y$germplasm_name <- factor(y$germplasm_name, levels = gens)
    y$study_name <- factor(y$study_name, levels = envs)
    
    # Format the genomic relationship matrix and environmental similarity matrix
    G <- as.matrix(current_data$subset_grm)[gens,gens]
    rownames(G) <- colnames(G) <- gens
    
    E <- as.matrix(current_data$subset_W)[envs,envs]
    rownames(E) <- colnames(E) <- envs
    
    # Remove test trial from training
    train_data <- y %>% filter(study_name != test_env)

    train_data$germplasm_name <- factor(train_data$germplasm_name, levels = rownames(G))
    train_data$study_name <- factor(train_data$study_name, levels = rownames(E))

    # Fit model
    multikernel_mod_fit <- mmer(
      grain_yield_kg_ha_co_321_0001218 ~ 1,
      random = ~ vs(germplasm_name, Gu = G) +
                vs(study_name, Gu = E),
      data = train_data
    )
  
  # Calculate model BLUPs
    mu <- multikernel_mod_fit$Beta$Estimate
    g_blup <- multikernel_mod_fit$U$`u:germplasm_name`$grain_yield_kg_ha_co_321_0001218
    e_blup <- multikernel_mod_fit$U$`u:study_name`$grain_yield_kg_ha_co_321_0001218
  
  # Predict test trial
    test_data <- y %>% filter(study_name == test_env)
    test_pred <- test_data %>%
      dplyr::mutate(
        g = g_blup[as.character(germplasm_name)],
        e = e_blup[as.character(study_name)],
        g = ifelse(is.na(g), 0, g),
        e = ifelse(is.na(e), 0, e),
        pred = mu + g + e
      )
    
  # Record Predictions
    predictions[[test_trial]] <- list(test_env = test_env, model = multikernel_mod_fit, predictions = test_pred)
  }
  return(predictions)
}
```

```{r run multikernel CV0 model}
multikernel_testset_predictions <- multi_GBLUP(training_data)
```


```{r Multi-Kernel Leave-One-Out Cross Validation function}
MK_LOO_CV <- function(training_data){
  
  CV_output <- vector("list", length(training_data))
  names(CV_output) <- names(training_data)
  
  for (test_trial in seq_along(training_data)){
    
    #Input data
    current_data <- training_data[[test_trial]]
    y <- current_data$combined_pheno |> 
      dplyr::filter(study_name != current_data$test_trial_metadata$study_name)
    
    # Assign fold IDs for each study
    study_assignments <- data.frame(study_name = unique(y$study_name), fold = seq(1:length( unique(y$study_name))))
    y <- y |> left_join(study_assignments, by = "study_name")
    
    # Get genotypes and environment information for all trials and set as factors
    gens <- unique(y$germplasm_name)
    envs <- unique(y$study_name)
    y$germplasm_name <- factor(y$germplasm_name, levels = gens)
    y$study_name <- factor(y$study_name, levels = envs)
    
    # Format the genomic relationship matrix and environmental similarity matrix
    G <- as.matrix(current_data$subset_grm)[gens,gens]
    rownames(G) <- colnames(G) <- gens
    
    E <- as.matrix(current_data$subset_W)[envs,envs]
    rownames(E) <- colnames(E) <- envs
    
    # Run CV
    cv_results <- list()

    for (fold_id in unique(y$fold)) {
  
      # Split data
      train_data <- dplyr::filter(y, fold != fold_id)
      test_data  <- dplyr::filter(y, fold == fold_id)
  
      # Subset G and E for training data
      train_gens <- unique(train_data$germplasm_name)
      train_envs <- unique(train_data$study_name)
  
      # Ensure factor levels match matrices
      train_data$germplasm_name <- factor(train_data$germplasm_name, levels = rownames(G))
      train_data$study_name <- factor(train_data$study_name, levels = rownames(E))
  
      # Fit model on training data
      cv_fit <- mmer(
        grain_yield_kg_ha_co_321_0001218 ~ 1,
        random = ~ vs(germplasm_name, Gu = G) +
               vs(study_name, Gu = E),
        data = train_data
      )
  
      # Extract BLUPs
      mu <- cv_fit$Beta$Estimate
      cv_g_blup <- cv_fit$U$`u:germplasm_name`$grain_yield_kg_ha_co_321_0001218
      cv_e_blup <- cv_fit$U$`u:study_name`$grain_yield_kg_ha_co_321_0001218
  
      # Predict test set
      test_pred <- test_data %>%
        dplyr::mutate(
          g = cv_g_blup[as.character(germplasm_name)],
          e = cv_e_blup[as.character(study_name)],
          pred = mu + g + e
        )
  
      cv_results[[fold_id]] <- test_pred %>% 
        dplyr::select(study_name, germplasm_name, grain_yield_kg_ha_co_321_0001218, pred)
    }
    # Combine predictions
    cv_df <- bind_rows(cv_results)

    # Accuracy metrics
    correlation <- if (sum(!is.na(cv_df$grain_yield_kg_ha_co_321_0001218) & !is.na(cv_df$pred)) > 0){
      cor(cv_df$grain_yield_kg_ha_co_321_0001218, cv_df$pred, use = "complete.obs")
    } else{
      NA_real_
    }
    rmse <- sqrt(mean((cv_df$grain_yield_kg_ha_co_321_0001218 - cv_df$pred)^2, na.rm = TRUE))
    
    within_env <- cv_df %>%
    dplyr::group_by(study_name) %>%
    dplyr::summarise(
      n_obs = sum(!is.na(`grain_yield_kg_ha_co_321_0001218`)),
      correlation = if (sum(!is.na(`grain_yield_kg_ha_co_321_0001218`) & !is.na(pred)) > 0){
        cor(`grain_yield_kg_ha_co_321_0001218`, pred, use = "complete.obs")
        } else {
          NA_real_
        },
      rmse = sqrt(mean((`grain_yield_kg_ha_co_321_0001218` - pred)^2, na.rm = TRUE))
    ) %>% arrange(desc(correlation))

    CV_output[[test_trial]] <- list(test_env = current_data$test_trial_metadata$study_name, CV_predictions = cv_df, correlation = correlation, rmse = rmse, within_env_acc = within_env)
  }
  
  return(CV_output)
  
} 
```

```{r run multikernel CV}
multikernel_CV <- MK_LOO_CV(training_data)
```

```{r prediction submissions}
for (test_trial in test_set){
  predictions <- multikernel_testset_predictions[[test_trial]]$predictions |>dplyr::select(germplasm_name,pred)
  colnames(predictions) <- c("germplasmName","prediction")
  write_csv(predictions, here::here("predictions",test_trial, "CV0_Predictions.csv"))
  
  trials <- data.frame(studyName=training_data[[test_trial]]$training_set_trials)
  write_csv(trials, here::here("predictions",test_trial, "CV0_Trials.csv"))
  
  accessions <- data.frame(germplasmName = training_data[[test_trial]]$training_accessions)
  write_csv(accessions, here::here("predictions",test_trial, "CV0_Accessions.csv"))
}
```


## CV 00

```{r Remove test accessions from training data}
CV00_training_data <- vector("list", length(training_data))
names(CV00_training_data) <- names(training_data)

for (test_trial in names(training_data)) {
  
  trial_data <- training_data[[test_trial]]
  
  # -------------------------------
  # Remove test-set accessions from training
  # -------------------------------
  training_accessions <- setdiff(trial_data$training_accessions, trial_data$test_accessions)
  
  # Keep only genotyped lines in GRM
  training_accessions <- intersect(training_accessions, rownames(trial_data$subset_grm))
  
  # -------------------------------
  # Subset phenotype for training only
  # -------------------------------
  subset_pheno <- trial_data$training_data %>%
    dplyr::filter(germplasm_name %in% training_accessions) %>%
    dplyr::select(study_db_id, study_name, germplasm_name, grain_yield_kg_ha_co_321_0001218)
  
  # -------------------------------
  # Keep test phenotypes (all NA)
  # -------------------------------
  test_pheno <- trial_data$test_pheno
  
  # -------------------------------
  # Combine phenotype
  # -------------------------------
  combined_pheno <- rbind(subset_pheno, test_pheno)
  
  # -------------------------------
  # Save CV00-ready data
  # -------------------------------
  CV00_training_data[[test_trial]] <- list(
    training_accessions = training_accessions,
    test_accessions = trial_data$test_accessions,
    subset_grm = trial_data$subset_grm,
    subset_W = trial_data$subset_W,
    combined_pheno = combined_pheno,
    test_trial_metadata = trial_data$test_trial_metadata
  )
}


```

```{r CV00 predictions}
CV00_multikernel_predictions <- multi_GBLUP(CV00_training_data)
```
```{r CV00 prediction submissions}
for (test_trial in test_set){
  predictions <- CV00_multikernel_predictions[[test_trial]]$predictions |>dplyr::select(germplasm_name,pred)
  colnames(predictions) <- c("germplasmName","prediction")
  write_csv(predictions, here::here("predictions",test_trial, "CV00_Predictions.csv"))
  
  trials <- data.frame(studyName=CV00_training_data[[test_trial]]$training_set_trials)
  write_csv(trials, here::here("predictions",test_trial, "CV00_Trials.csv"))
  
  accessions <- data.frame(germplasmName = CV00_training_data[[test_trial]]$training_accessions)
  write_csv(accessions, here::here("predictions",test_trial, "CV00_Accessions.csv"))
}
```
