---
title: "Pull genotype data from T3"
author: "Prediction5 Team"
output: html_document
date: '`r format(Sys.time(), "%d %B, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

# Libraries

```{r load libraries}
suppressPackageStartupMessages(library(here)) # root directory
suppressPackageStartupMessages(library(data.table)) # load data
suppressPackageStartupMessages(library(apercu)) # ap
suppressPackageStartupMessages(library(dplyr)) # formatting
suppressPackageStartupMessages(library(purrr)) # handle lists
## devtools::install_github("TriticeaeToolbox/BrAPI.R")
suppressPackageStartupMessages(library(BrAPI)) # T3 connection
suppressPackageStartupMessages(library(T3BrapiHelpers)) # T3 helper
suppressPackageStartupMessages(library(DT)) # print table
suppressPackageStartupMessages(library(ggplot2)) # plot
suppressPackageStartupMessages(library(vcfR)) # load vcf
suppressPackageStartupMessages(library(progress)) # progress bar

```

# Set up root directory

```{r current dir}
here::i_am("analysis/combine_GRM.Rmd")
source(here("code","useful_functions.R"))
```

```{r}
target.tri <- c("AWY1_DVPWA_2024","TCAP_2025_MANKS","25_Big6_SVREC_SVREC",
                "OHRWW_2025_SPO","CornellMaster_2025_McGowan","24Crk_AY2-3",
                "2025_AYT_Aurora","YT_Urb_25","STP1_2025_MCG")
```

# Convert vcf files to GRM matrices

For non-focal accessions

```{r}
# tri=target.tri[9]
# for(tri in target.tri){
# 
#   write(paste0("Processing trial: ", tri), stdout())
#   files <- list.files(here("data","genotypic",tri,"nonFocal"))
#   files <- grep(files, pattern=".vcf.gz", value=TRUE)
#   files <- grep(files, pattern="_formatted|_imputed", value=TRUE, invert=TRUE)## keep only the unformatted vcf files
#   for(file in files){
#     
#     vcf.p2f <- paste0(here("data","genotypic",tri,"nonFocal"),"/",file)
# 
#     ## get filename without extension
#     file_noext <- gsub(pattern=".vcf.gz","",file)
#     
#     p2f.export.vcf <- paste0(here("data","genotypic",tri,"nonFocal"),"/",file_noext,"_formatted.vcf.gz")
#     ## use custom function from useful_functions.R
#     geno <- format_curate_vcf(vcf.p2f=vcf.p2f,
#                               p2f.export.vcf=p2f.export.vcf,
#                               remove.chrUkn=FALSE,
#                               check.mrk.dups=TRUE,
#                               remove.nonPolyMrk=FALSE,
#                               tresh.heterozygous=NULL,
#                               check.geno.dups=TRUE,
#                               thresh.NA.ind=0.5,
#                               thresh.NA.mrk=0.5,
#                               format.names=FALSE,
#                               imputation="Beagle",
#                               p2f.beagle=here("code","beagle.27Feb25.75f.jar"),
#                               thresh.MAF=0.01,
#                               verbose=1)
# 
#     ## export numeric imputed geno file
#     p2f.num <- paste0(here("data","genotypic",tri,"nonFocal"),"/",file_noext,"_numeric.rds")
#     saveRDS(geno, file=p2f.num)
# 
#     ## calculate the genomic relationship matrix
#     grm <- rrBLUP::A.mat(geno-1)
#     ## export grm
#     p2f.grm <- paste0(here("data","genotypic",tri,"nonFocal"),"/",file_noext,"_GRM.rds")
#     saveRDS(grm, file=p2f.grm)
#   }
# }
```

# Combine GRM matrices for each program

For AWY1, same vcf files were downloaded twice, copy the focal_accessions GRM to the combined_GRM. For Cornell, two files were identical but completely disjoint from the focalAccessions. Copy the focalAccessions_GRM to the combined_GRM.

```{r combine-grm}
pbm <- c("CornellMaster_2025_McGowan", "AWY1_DVPWA_2024")
rerun=TRUE
for(tri in setdiff(target.tri, pbm)){
  p2f.export <- paste0(here("data","genotypic",tri,"clean"),"/",tri,"_combined_GRM.rds")
  if(!file.exists(p2f.export)|rerun){
    
    write(paste0("Processing trial: ", tri), stdout())
    grmList <- list()
    
    ## Step 1: convcert the focal accessions genomic matrix to a GRM
    p2f <- paste0(here("data","genotypic",tri,"clean"),"/",tri,"_Predictathon_genotypes_curated_numeric.rds")
    genoFoc <- readRDS(p2f)
    grmList[[1]] <- rrBLUP::A.mat(genoFoc-1)
    ## export focal GRM
    p2f <- paste0(here("data","genotypic",tri,"clean"),"/",tri,"_focalAccessions_GRM.rds")
    saveRDS(grmList[[1]], file=p2f)
    
    ## Step 2: load the non focal GRM matrices
    files <- list.files(here("data","genotypic",tri,"nonFocal"))
    ## keep only the GRM files
    files <- grep("_GRM.rds", files, value = TRUE)
    grmList <- c(grmList, lapply(files, function(f){
      readRDS(paste0(here("data","genotypic",tri,"nonFocal"),"/",f))
      # geno = readRDS(p2f)
      # return(rrBLUP::A.mat(geno-1))
    }))
    genoNames <- map(grmList,rownames)
    ap(genoNames)
    
    ## if same genotypes, skip
    if(length(genoNames) > 1){
      ggVennDiagram::ggVennDiagram(genoNames, label_alpha = 0, label_size = 4) 
      if(all(genoNames[[1]] %in% genoNames[[2]]) & all(genoNames[[2]] %in% genoNames[[1]])){
        genoNames[[2]] <- NULL
        next
      } else {
        
        ## combine genomic relationship matrices
        var_indices <- list()
        for(i in seq_along(grmList)){
          var_indices[[i]] <- match(rownames(grmList[[i]]), unique(unlist(genoNames)))-1
        }
        
        # combine GRM matrices
        comb_grm <- covariance_combiner(
          partial_covs=grmList,
          var_indices=var_indices,
          degrees_freedom=rep(100, length(grmList)),
          calc_sampling_cov = FALSE
        )
        dim(comb_grm$psi)
        grm_combined <- comb_grm$psi
        rownames(grm_combined) <- colnames(grm_combined) <- unique(unlist(genoNames))
        ## export combined GRM
        p2f.export <- paste0(here("data","genotypic",tri,"clean"),"/",tri,"_combined_GRM.rds")
        saveRDS(grm_combined, file=p2f.export)
      }
    }
  }
}
```

## Combine Focal and Combined GRMs

```{r}
# Define the root directory where all your trial folders are located
base_dir <- here::here("scripts", "Prediction5", "data", "genotypic")
trial_folders <- list.dirs(base_dir, full.names = TRUE, recursive = FALSE)
combined_grms <- list()

for (folder in trial_folders) {
  trial_name <- basename(folder)
  write(paste0("\n=== Processing: ", trial_name, " ==="), stdout())
  
  clean_dir <- file.path(folder, "clean")
  if (!dir.exists(clean_dir)) {
    write("   -> SKIPPED: No 'clean' folder.", stderr())
    next
  }
  
  # ====================================================================
  # NEW: Only pull files that end specifically in "GRM.rds"
  rds_files <- list.files(clean_dir, pattern = "GRM\\.rds$", full.names = TRUE, ignore.case = TRUE)
  # ====================================================================
  
  write(paste0("   -> Found ", length(rds_files), " GRM.rds files."), stdout())
  
  if (length(rds_files) < 2) {
    write("   -> SKIPPED: Needs at least 2 GRM files to combine.", stderr())
    next
  }
  
  # Load and sanitize the matrices
  partial_covs <- lapply(rds_files, function(f) {
    mat <- readRDS(f)
    
    # Sanitize: Drop extra text columns if it loaded as a data frame
    if (is.data.frame(mat) && (is.character(mat[[1]]) || is.factor(mat[[1]]))) {
      rownames(mat) <- as.character(mat[[1]])
      mat <- mat[, -1, drop = FALSE] 
    }
    
    mat <- as.matrix(mat)
    
    # Sanitize: Force numeric if it accidentally became character
    if (typeof(mat) == "character") {
      rnames <- rownames(mat)
      cnames <- colnames(mat)
      mat <- matrix(as.numeric(mat), nrow = nrow(mat), ncol = ncol(mat))
      rownames(mat) <- rnames
      colnames(mat) <- cnames
    }
    
    # Sanitize: Remove duplicate row names
    if (any(duplicated(rownames(mat)))) {
      mat <- mat[!duplicated(rownames(mat)), !duplicated(colnames(mat))]
    }
    
    return(mat)
  })
  
  # Create the Master Dictionary
  all_gids <- unique(unlist(lapply(partial_covs, rownames)))
  all_gids <- sort(all_gids) 
  
  # Map 0-based indices for the EM function
  var_indices <- lapply(partial_covs, function(mat) {
    return(match(rownames(mat), all_gids) - 1)
  })
  
  write(paste0("   -> Combining ", length(partial_covs), " matrices (Total lines: ", length(all_gids), ")..."), stdout())
  
  combined_result <- tryCatch({
    covariance_combiner(
      partial_covs = partial_covs, var_indices = var_indices,
      degrees_freedom = NULL, max_iter = 100, tol = 1e-6,
      track_loglik = FALSE, calc_sampling_cov = FALSE
    )
  }, error = function(e) {
    write(paste0("   -> ERROR IN MATH: ", e$message), stderr())
    return(NULL)
  })
  
  if (is.null(combined_result)) next
  
  # Reformat the final combined matrix
  combined_matrix <- combined_result$psi
  rownames(combined_matrix) <- all_gids
  colnames(combined_matrix) <- all_gids
  combined_matrix <- (combined_matrix + t(combined_matrix)) / 2 # Force Symmetry
  
  combined_grms[[trial_name]] <- combined_matrix
  
  save_path <- file.path(clean_dir, paste0(trial_name, "_All_GRM.rds"))
  saveRDS(combined_matrix, save_path)
  write(paste0("   -> SUCCESS! Saved combined matrix to: ", save_path), stdout())
}
```
