---
title: "Apply genomic prediction"
author: "Prediction5 Team"
output: html_document
df_print: kable
date: '`r format(Sys.time(), "%d %B, %Y")`'
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE,
                      engine.opts = list(bash = "-l"))
```

# Libraries

```{r load libraries}
suppressPackageStartupMessages(library(here)) # root directory
suppressPackageStartupMessages(library(data.table)) # load data
suppressPackageStartupMessages(library(apercu)) # ap
suppressPackageStartupMessages(library(rrBLUP)) # genomic selection
suppressPackageStartupMessages(library(dplyr)) # formatting
suppressPackageStartupMessages(library(ggplot2)) # plot
suppressPackageStartupMessages(library(paletteer)) # plot palette
suppressPackageStartupMessages(library(janitor)) # formating
suppressPackageStartupMessages(library(BGLR)) # Bayes GP
suppressPackageStartupMessages(library(foreach)) # for parallelization
suppressPackageStartupMessages(library(snow)) # for creating cluster
suppressPackageStartupMessages(library(doSNOW)) # for creating cluster
suppressPackageStartupMessages(library(purrr)) # list handling
suppressPackageStartupMessages(library(bigstatsr)) # ggplot theme
suppressPackageStartupMessages(library(ggbeeswarm)) # ggplot geom_quasirandom
library(stringr)
library(tidyverse)

nb.cores <- 6
```

# Set up root directory

```{r current dir}
here::i_am("analysis/genom_pred.Rmd")
source(here("code","useful_functions.R")) # getFolds function
```

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:

```{r time_0}
t0 <- proc.time()
```

```{r}
target.pred <- c("AWY1_DVPWA_2024","TCAP_2025_MANKS","25_Big6_SVREC_SVREC",
                 "OHRWW_2025_SPO","CornellMaster_2025_McGowan","24Crk_AY2-3",
                 "2025_AYT_Aurora","YT_Urb_25","STP1_2025_MCG")
tri <- "2025_AYT_Aurora"
```

# Load data

## Genomic data

```{r load_geno}
## option 1: load combined GRM for each trial
#p2fs <- here("data","genotypic", target.pred,"clean",
#             paste0(target.pred,"_All_GRM.rds"))


#file.mtime(p2fs)
#grmList <- lapply(p2fs, readRDS)
combined_trials <- c("25_Big6_SVREC_SVREC", "YT_Urb_25")

file_endings <- ifelse(target.pred %in% combined_trials, 
                       "_combined_GRM.rds", 
                       "_All_GRM.rds")

# Build the exact file paths using those dynamic endings
p2fs <- here("data", "genotypic", target.pred, "clean", 
             paste0(target.pred, file_endings))

# Check modification times (from your original code)
file.mtime(p2fs)

# Load them all into your list
grmList <- lapply(p2fs, readRDS)

names(grmList) <- target.pred

## option 2: load genotypic data from Predictathon downloaded vcfs
# p2fs <- here("data","genotypic", target.pred,"clean",
#              paste0(target.pred,"_Predictathon_genotypes_curated_numeric.rds"))
# file.mtime(p2fs)
# genoList <- lapply(p2fs, readRDS)
# names(genoList) <- target.pred
# ap(genoList)
# sapply(genoList, dim)
# grmList <- lapply(genoList, function(x) x=rrBLUP::A.mat(x-1))

ap(grmList)
```

Small number of genotypes for "AWY1_DVPWA_2024"

## Phenotypic data

Phenotypic data formatted and generated in `phenotypic_data.Rmd`.

```{r load_pheno_trials}
p2f <- here("data","phenotypic","allAccessions_yield_raw.csv")
p2f <- here("data","phenotypic","allTrials_pheno_genoSelected.csv")

file.mtime(p2f)
pheno_init <- fread(p2f)
dim(pheno_init)
head(pheno_init)
str(pheno_init)
pheno <- pheno_init |> 
  rename(GID=germplasm_name, YLD=value, PROG=program_name, LOC=location_name, YEAR=year) |>
  mutate(STATE = str_extract(LOC, "[A-Z]{2}$")) |> 
  mutate(LOC = gsub(", [A-Z]{2}", "", LOC)) |> 
  mutate(ENV = paste0(LOC, "_", YEAR)) |>
  dplyr::select(GID, PROG, LOC, STATE, YEAR, ENV, YLD,study_name) |> 
  group_by(GID, LOC, STATE, YEAR, ENV,study_name) |>
  summarise(YLD = mean(YLD, na.rm=TRUE), .groups = "drop")
head(pheno)

pheno <- pheno %>% filter(YEAR >= 2020)
```

## Common genotypes

```{r}
grmList.cmon <- purrr::map(grmList, \(x){
  int <- intersect(rownames(x), unique(pheno$GID))
  return(x[int,int])
})
lapply(grmList.cmon, nrow)
```

# GBLUP

## CV0

Cross validation by program. Predict unknown trial with known genotypes. Steps:

-   Subset the lines both genotyped and phenotyped for each target trial
-   Predict each environment in turn using the other trials as training set.

```{r gblup_cv0}
res.cv0 <- expand.grid(method="GBLUP", scenario="CV0",
                       trial=target.pred,  corP=NA,
                       wEnv.corP=NA,nb.genos=NA, nb.envs=NA)
obs.pred.all <- data.frame()

for(tri in target.pred){
  write(paste0("Processing trial: ", tri,"\n"), stderr())
  
  ## subset GID with geno and pheno data
  gid.geno <- rownames(grmList[[tri]])
  gid <- intersect(gid.geno, pheno$GID)
  length(gid)
  if(length(gid) == 0){
    write(paste0("No genotypes with both geno and pheno data for trial ", tri), stdout())
    
  }else {
    ## subset pheno and geno from the program to predict
    geno_sub <- grmList[[tri]][gid,gid]
    pheno_sub <- pheno |> filter(GID %in% gid)
    ## select environments with more than 30 obs
    envs <- pheno_sub %>%
      group_by(ENV) %>% summarize(n=n(),.groups="drop") %>%
      filter(n > 30) %>% pull(ENV)
    #pheno_sub <- pheno_sub %>% filter(ENV %in% envs)
    if(length(envs) == 0){
      write(paste0("No environment with more than 30 observations for trial ", tri), stdout())
      next
    } else {
      df_pred_e <- data.frame()
      ## for loop over environments to predict each environment in turn
      for(e in envs){
        ## set yield data for the test environment to NA
        pheno_train <- pheno_sub |>
          mutate(YLD=ifelse(ENV %in% e, NA, YLD)) %>%
          as.data.frame()
        
        ## fit GBLUP model
        ### add fixed effects of location and year instead of env?
        fit <- rrBLUP::kin.blup(data=pheno_train, geno="GID",pheno="YLD",
                                K=geno_sub,
                                fixed=c("LOC","YEAR"),#"ENV", ## add env as fixed effect
                                n.core=nb.cores)
        ## contains the predicted values for all genotypes, including those with observed phenotypes in the training set
        #stopifnot(identical(names(fit$pred), pheno_train$GID))
        ## for rrBLUP, predicted values averaged over fixed effects
        pred.all <- data.frame(ENV=e,
                               GID=names(fit$pred), pred=c(fit$pred))
        ## export predicted values for the test environment
        df_pred_e <- rbind(df_pred_e,filter(pred.all, ENV %in% e))
        
      }
      ## add observed values
      df_pred_e <- df_pred_e |> left_join(pheno_sub, by=c("GID","ENV")) %>% 
        mutate(target.trial=tri,
               scenario="CV0",
               GP.method="GBLUP")
      obs.pred.all <- rbind(obs.pred.all, df_pred_e)
      
      plot(ggplot(df_pred_e, aes(x=YLD, y=pred, color=ENV)) +
             geom_point(alpha=0.6, size=1) +
             geom_smooth(method="lm", se=FALSE)+
             geom_rug(sides="bl")+
             #paletteer::scale_color_paletteer_d("ggthemes::Jewel_Bright") +
             ggpubr::stat_cor(method="pearson", size=3)+
             geom_abline(intercept=0, slope=1, color="grey", linetype="dashed")+
             ggtitle(paste0("Predicted vs. observed values for trial ", tri),
                     subtitle=paste0("GBLUP genomic selection method"))+
             theme_bigstatsr(size.rel=0.8))
      
      ## calculate PA in each environment
      pa.within.env <- df_pred_e %>% group_by(ENV) %>%
        summarize(pa=cor(YLD, pred, use="pairwise.complete.obs")) %>%
        as.data.frame()
      ## calculate correlation across environments
      pa <- round(cor(df_pred_e$YLD, df_pred_e$pred, use="pairwise.complete.obs"),3)
      idx <- which(res.cv0$trial == tri)
      res.cv0$corP[idx] <- round(pa,3)
      res.cv0$wEnv.corP[idx] <- round(mean(pa.within.env$pa, na.rm=TRUE),3)
      res.cv0$nb.envs[idx] <- length(envs)
      ## add training/validation set size of the last fold
      res.cv0$nb.genos[idx] <- length(unique(pheno_train$GID[!is.na(pheno_train$YLD)]))
    }
  }
}
print_table(res.cv0)
write.table(res.cv0, here("output","GP_GBLUP_CV0_combined-GRM_common-pheno.tsv"),
            sep="\t", row.names=FALSE)

saveRDS(obs.pred.all, file=here("output","GP_GBLUP_CV0_combined-GRM_common-pheno_obs-pred.rds"))
```

## CV0 w/ Environmental Relationship Matrix

### Preparing He

```{r}
trial_env <- read.csv(here("analysis", "environmental_covariates.csv"))
trial_env <- trial_env %>% filter(yday >= 60 & yday <= 245)
trial_env <- trial_env %>% filter(YEAR >= 2020)
env_data <- trial_env[,c(11:17)]
ec_scaled <- scale(env_data) # remove Env name column before scaling
ec_scaled <- as.data.frame(ec_scaled)
ec_scaled$ENV <- trial_env$ENV
ec_scaled$DATE <- trial_env$yday
ec_scaled$LOC <- trial_env$location

ec_scaled <- unique(ec_scaled)

ec_scaled <- ec_scaled %>% pivot_wider(
  names_from = DATE,
  values_from = c("dayl..s.", "prcp..mm.day.", "srad..W.m.2.", "swe..kg.m.2.", "tmax..deg.c.", "tmin..deg.c.", "vp..Pa.")
)

He <- ec_scaled
He<- He[,-c(1,2)]
rownames(He) = ec_scaled$ENV

He <- cor(t(He))
```

### CV0 Predictions Sommer

```{r}
library(dplyr)
library(sommer)

target_programs <- c("AWY1_DVPWA_2024")

# Prepare an empty dataframe to store the CV0 predictive abilities
cv_results <- data.frame(Trial = character(), Masked_Env = character(), 
                         Pearson_r = numeric(), Valid_N = numeric(), stringsAsFactors = FALSE)

for(tri in target_programs){
  write(paste0("\n======================================================\n",
               "Starting CV0 for Breeding Program/Trial: ", tri,
               "\n======================================================"), stderr())
  
  # 1. Isolate Genetics and Phenotypes for this program
  if(!tri %in% names(grmList)){
    write(paste0("   Error: ", tri, " not found in grmList! Skipping."), stderr())
    next
  }
  
  gid.geno <- rownames(grmList[[tri]])
  geno_sub <- grmList[[tri]]
  pheno_sub <- pheno %>% dplyr::filter(GID %in% gid.geno)
  
  # =====================================================================
  # THE RAM SAVER (Historical Filter)
  # Drop tiny environments to drastically reduce the Kronecker matrix size
  robust_envs <- pheno_sub %>%
    dplyr::group_by(ENV) %>%
    dplyr::summarize(n = n(), .groups = "drop") %>%
    dplyr::filter(n >= 25) %>% # Adjust this number up if you still run out of RAM!
    dplyr::pull(ENV)
  
  pheno_sub <- pheno_sub %>% dplyr::filter(ENV %in% robust_envs)
  # =====================================================================
  
  # Identify valid environments that also exist in the He matrix
  envs <- unique(pheno_sub$ENV)
  envs <- intersect(envs, rownames(He))
  
  if(length(envs) < 2){
    write(paste0("   Skipping ", tri, ": Not enough valid environments left for cross-validation."), stderr())
    next
  }
  
  # Begin Leave-One-Environment-Out Loop
  for(e in envs){
    gc() # Force clear RAM
    write(paste0("   Masking and predicting environment: ", e), stderr())
    
    # 2. Mask the target environment
    pheno_train <- pheno_sub %>% dplyr::filter(ENV %in% envs)
    
    # Save the true values for correlation later
    true_pheno <- pheno_train %>% dplyr::filter(ENV == e & !is.na(YLD))
    
    # Mask YLD for the target environment
    pheno_train$YLD[pheno_train$ENV == e] <- NA
    
    # 3. Safety Check: Ensure we have training data left
    if(sum(!is.na(pheno_train$YLD)) < 20){
      write("      Skipping: Not enough training data after masking.", stderr())
      next
    }
    
    # 4. Build Localized Matrices
    all_envs <- unique(pheno_train$ENV)
    all_gids <- unique(pheno_train$GID)
    
    He_sub <- He[all_envs, all_envs]
    Kg_sub <- geno_sub[all_gids, all_gids]
    
    K_ge <- kronecker(He_sub, Kg_sub)
    ge_names <- paste(rep(rownames(He_sub), each = length(all_gids)),
                      rep(rownames(Kg_sub), times = length(rownames(He_sub))), sep = ":::")
    rownames(K_ge) <- colnames(K_ge) <- ge_names
    
    # Lock GE_ID factor levels to the K_ge matrix dimensions
    pheno_train <- pheno_train %>% 
      dplyr::mutate(GE_ID = factor(paste(ENV, GID, sep = ":::"), levels = rownames(K_ge)))
    
    # 5. Run Model (Main Effect + GxE)
    fit <- tryCatch({
      mmer(fixed = YLD ~ LOC + YEAR, 
           random = ~ vsr(GID, Gu = Kg_sub) + vsr(GE_ID, Gu = K_ge), 
           rcov = ~ units, 
           data = pheno_train,
           verbose = FALSE)
    }, error = function(err){
      write(paste0("      Model crashed for ", e, ": ", err$message), stderr())
      return(NULL)
    })
    
    if(is.null(fit)) next
    
    # 6. Manual Extraction of Main Effect + GxE
    # Extract Main Genotype Effect
    blups_main <- fit$U$`u:GID`[[1]]
    if(is.null(blups_main)) blups_main <- fit$U[[1]][[1]]
    
    main_df <- data.frame(
      GID = gsub("^GID", "", names(blups_main)),
      Main_Effect = as.numeric(blups_main)
    )
    
    # Extract GxE Effect
    blups_gxe <- fit$U$`u:GE_ID`[[1]]
    if(is.null(blups_gxe)) blups_gxe <- fit$U[[2]][[1]]
    
    gxe_df <- data.frame(
      ENV = sub(":::.*", "", gsub("^GE_ID", "", names(blups_gxe))),
      GID = sub(".*:::", "", gsub("^GE_ID", "", names(blups_gxe))),
      GxE_Effect = as.numeric(blups_gxe)
    )
    
    # Combine and add intercept
    pop_mean <- mean(pheno_clean$YLD, na.rm = TRUE)
    
    predictions <- gxe_df %>%
      dplyr::filter(ENV == e) %>% # Only keep predictions for the masked environment
      dplyr::left_join(main_df, by = "GID") %>%
      dplyr::mutate(Predicted_YLD = pop_mean + Main_Effect + GxE_Effect)
    
    # 7. Calculate Predictive Ability (Pearson Correlation)
    eval_df <- true_pheno %>%
      dplyr::left_join(predictions, by = c("ENV", "GID")) %>%
      dplyr::filter(!is.na(Predicted_YLD) & !is.na(YLD))
    
    if(nrow(eval_df) > 5){
      r_val <- cor(eval_df$YLD, eval_df$Predicted_YLD, use = "complete.obs")
      write(paste0("      Accuracy (r): ", round(r_val, 3), " (n=", nrow(eval_df), ")"), stdout())
      
      cv_results <- rbind(cv_results, data.frame(
        Trial = tri, Masked_Env = e, Pearson_r = r_val, Valid_N = nrow(eval_df)
      ))
    }
    
    # Memory cleanup
    rm(K_ge, fit, He_sub, Kg_sub, pheno_train, eval_df, predictions)
  }
}

write.csv(cv_results, "CV0_Predictive_Abilities_Updated.csv", row.names = FALSE)
write("CV0 Complete! Results saved.", stdout())
```

### CV0 Final Predictions

```{r}
library(dplyr)
library(readr)
library(sommer)

# 1. Load the target accessions
target_data <- read_csv(here("data","acc_for_prediction.csv"))

target_pheno <- target_data %>%
  dplyr::rename(GID = acc_name,
                LOC = location,
                YEAR = year) %>%  
  dplyr::mutate(YLD = NA) %>%
  dplyr::select(trial, ENV, GID, LOC, YEAR, YLD)

all_final_predictions <- data.frame()
target_trials <- unique(target_pheno$trial)

library(parallel)
library(dplyr)

# 1. Define the number of safe cores to use
safe_cores <- 10

write(paste0("Starting parallel processing using ", safe_cores, " cores..."), stderr())

# 2. Convert your loop into a function
process_trial <- function(tri) {
  # Force garbage collection inside the worker node
  gc() 
  
  target_sub <- target_pheno %>% dplyr::filter(trial == tri)
  
  if(!tri %in% names(grmList)){
    return(NULL) # Replaces 'next' in a function
  }
  
  gid.geno <- rownames(grmList[[tri]])
  geno_sub <- grmList[[tri]]
  
  target_sub <- target_sub %>% dplyr::filter(GID %in% gid.geno)
  if(nrow(target_sub) == 0){
    return(NULL)
  }
  
  pheno_sub <- pheno %>% dplyr::filter(GID %in% gid.geno)
  
  pheno_clean <- pheno_sub %>%
    dplyr::anti_join(target_sub, by = c("ENV", "GID"))
  
  # The RAM Saver Filter
  robust_historical_envs <- pheno_clean %>%
    dplyr::group_by(ENV) %>%
    dplyr::summarize(n = n(), .groups = "drop") %>%
    dplyr::filter(n >= 30) %>%  
    dplyr::pull(ENV)
  
  # === NEW: Define your diverse trial here ===
  rescue_env <- "Neoga_2022" 
  # ===========================================
  
  target_envs <- unique(target_sub$ENV)
  
  # Force the rescue environment into the allowed list so it doesn't get dropped
  envs_to_keep <- unique(c(target_envs, robust_historical_envs, rescue_env))
  
  pheno_clean <- pheno_clean %>% dplyr::filter(ENV %in% envs_to_keep)
  
  # Target-specific filter for YT_Urb_25
  if(tri == "YT_Urb_25"){
    pheno_clean <- pheno_clean %>%
      dplyr::mutate(numeric_year = as.numeric(as.character(YEAR)))
    
    recent_years <- sort(unique(pheno_clean$numeric_year), decreasing = TRUE)[1:3]
    
    # Keep the last 2 years OR the specific rescue environment
    pheno_clean <- pheno_clean %>%
      dplyr::filter(numeric_year %in% recent_years | ENV == rescue_env) %>%
      dplyr::select(-numeric_year)
  }
  
  pheno_combined <- dplyr::bind_rows(pheno_clean, 
                                     target_sub %>% dplyr::select(-trial))
  
  missing_envs <- setdiff(target_envs, rownames(He))
  if(length(missing_envs) > 0){
    return(NULL)
  }
  
  valid_envs <- intersect(unique(pheno_combined$ENV), rownames(He))
  pheno_combined <- pheno_combined %>% dplyr::filter(ENV %in% valid_envs)
  all_gids <- unique(pheno_combined$GID) 
  
  # Build Matrices
  He_sub <- He[valid_envs, valid_envs]
  Kg_sub <- geno_sub[all_gids, all_gids]
  
  K_ge <- kronecker(He_sub, Kg_sub)
  ge_names <- paste(rep(rownames(He_sub), each = length(all_gids)),
                    rep(rownames(Kg_sub), times = length(rownames(He_sub))), sep = ":::")
  rownames(K_ge) <- colnames(K_ge) <- ge_names
  
  pheno_combined <- pheno_combined %>% 
    dplyr::mutate(GE_ID = factor(paste(ENV, GID, sep = ":::"), levels = rownames(K_ge)))
  
  # Run the Model 
  fit <- tryCatch({
    sommer::mmer(fixed = YLD ~ LOC + YEAR, 
                 random = ~ sommer::vsr(GID, Gu = Kg_sub) + sommer::vsr(GE_ID, Gu = K_ge), 
                 rcov = ~ units, 
                 data = pheno_combined,
                 verbose = FALSE) 
  }, error = function(e){
    return(NULL)
  })
  
  if(is.null(fit)) return(NULL)
  
  # Extract Main Genotype Effect
  blups_main <- fit$U$`u:GID`[[1]]
  if(is.null(blups_main)) blups_main <- fit$U[[1]][[1]]
  
  main_df <- data.frame(
    GID = gsub("^GID", "", names(blups_main)),
    Main_Effect = as.numeric(blups_main)
  )
  
  # Extract GxE Effect
  blups_gxe <- fit$U$`u:GE_ID`[[1]]
  if(is.null(blups_gxe)) blups_gxe <- fit$U[[2]][[1]]
  
  gxe_df <- data.frame(
    GE_ID = names(blups_gxe),
    GxE_Effect = as.numeric(blups_gxe)
  ) %>%
    dplyr::mutate(
      ENV = sub(":::.*", "", GE_ID),
      GID = sub(".*:::", "", GE_ID)
    )
  
  pop_mean <- mean(pheno_clean$YLD, na.rm = TRUE)
  
  # Combine
  trial_predictions <- gxe_df %>%
    dplyr::left_join(main_df, by = "GID") %>%
    dplyr::mutate(Predicted_YLD = pop_mean + Main_Effect + GxE_Effect,
                  standard.error = NA) %>%
    dplyr::select(ENV, GID, Predicted_YLD, standard.error)
  
  trial_export <- target_sub %>%
    dplyr::left_join(trial_predictions, by = c("ENV", "GID"))
  
  # Instead of binding rows immediately, we return the dataframe
  return(trial_export)
}

# 3. Execute the function across all trials using 10 cores simultaneously
results_list <- mclapply(target_trials, process_trial, mc.cores = safe_cores)

# 4. Bind all the successful parallel runs into one final dataframe
all_final_predictions <- dplyr::bind_rows(results_list)

write("Parallel processing complete!", stderr())
# Format and Save
write.csv(all_final_predictions, "CV0_Final_Yield_Predictions.csv", row.names = FALSE)
write("\nAll predictions complete and saved", stdout())
```

## CV00 w/ Environmental Relationship Matrix

### CV00 Predictions Sommer

```{r}
library(dplyr)
library(sommer)

# Define target programs (or use names(grmList))
target_programs <- c("STP1_2025_MCG") 

cv_results <- data.frame(Trial = character(), Target_Env = character(), Fold = numeric(),
                         Pearson_r = numeric(), Valid_N = numeric(), stringsAsFactors = FALSE)

for(tri in target_programs){
  gc()
  write(paste0("\n======================================================\n",
               "Starting Strict CV00 for: ", tri, "\n======================================================"), stderr())
  
  if(!tri %in% names(grmList)) next
  
  gid.geno <- rownames(grmList[[tri]])
  geno_sub <- grmList[[tri]]
  pheno_sub <- pheno %>% dplyr::filter(GID %in% gid.geno)
  
  # RAM SAVER
  robust_envs <- pheno_sub %>% 
    dplyr::group_by(ENV) %>% 
    dplyr::summarize(n = n(), .groups = "drop") %>% 
    dplyr::filter(n >= 25) %>% 
    dplyr::pull(ENV)
  
  pheno_sub <- pheno_sub %>% dplyr::filter(ENV %in% robust_envs)
  envs <- intersect(unique(pheno_sub$ENV), rownames(He))
  
  if(length(envs) < 2) next
  
  # Loop through environments
  for(e in envs){
    env_data <- pheno_sub %>% dplyr::filter(ENV == e & !is.na(YLD))
    gids_in_env <- unique(env_data$GID)
    
    if(length(gids_in_env) < 15) next # Skip if too few genotypes
    
    # Create 5 Folds for the Genotypes
    set.seed(42) 
    folds <- sample(rep(1:5, length.out = length(gids_in_env)))
    split_gids <- split(gids_in_env, folds)
    
    for(f in 1:5){
      gc()
      focal_gids <- split_gids[[f]]
      write(paste0("   Env: ", e, " | Fold: ", f, " | Masking ", length(focal_gids), " focal lines..."), stderr())
      
      # 1. STRICT CV00 PURGE:
      # - Remove focal lines completely (Untested Lines)
      # - Remove the target environment completely (Untested Environment)
      pheno_train <- pheno_sub %>% 
        dplyr::filter(!(GID %in% focal_gids)) %>%
        dplyr::filter(ENV != e) 
      
      # 2. Add focal lines back ONLY in the target environment as NA
      target_na <- env_data %>% 
        dplyr::filter(GID %in% focal_gids) %>% 
        dplyr::mutate(YLD = NA)
      
      pheno_combined <- dplyr::bind_rows(pheno_train, target_na)
      
      # 3. Build Matrices
      all_envs <- unique(pheno_combined$ENV)
      all_gids <- unique(pheno_combined$GID)
      
      He_sub <- He[all_envs, all_envs]
      Kg_sub <- geno_sub[all_gids, all_gids]
      
      # FORCE SYMMETRY FOR C++
      He_sub <- (He_sub + t(He_sub)) / 2
      Kg_sub <- (Kg_sub + t(Kg_sub)) / 2
      
      K_ge <- kronecker(He_sub, Kg_sub)
      ge_names <- paste(rep(rownames(He_sub), each = length(all_gids)),
                        rep(rownames(Kg_sub), times = length(rownames(He_sub))), sep = ":::")
      rownames(K_ge) <- colnames(K_ge) <- ge_names
      
      pheno_combined <- pheno_combined %>% 
        dplyr::mutate(GE_ID = factor(paste(ENV, GID, sep = ":::"), levels = rownames(K_ge)))
      
      # 4. Run Model
      fit <- tryCatch({
        mmer(fixed = YLD ~ LOC + YEAR, 
             random = ~ vsr(GID, Gu = Kg_sub) + vsr(GE_ID, Gu = K_ge), 
             rcov = ~ units, data = pheno_combined, verbose = FALSE)
      }, error = function(err) return(NULL))
      
      if(is.null(fit)) next
      
      # 5. Extract BLUPs
      blups_main <- fit$U$`u:GID`[[1]]
      if(is.null(blups_main)) blups_main <- fit$U[[1]][[1]]
      main_df <- data.frame(GID = gsub("^GID", "", names(blups_main)), Main_Effect = as.numeric(blups_main))
      
      blups_gxe <- fit$U$`u:GE_ID`[[1]]
      if(is.null(blups_gxe)) blups_gxe <- fit$U[[2]][[1]]
      gxe_df <- data.frame(
        ENV = sub(":::.*", "", gsub("^GE_ID", "", names(blups_gxe))),
        GID = sub(".*:::", "", gsub("^GE_ID", "", names(blups_gxe))),
        GxE_Effect = as.numeric(blups_gxe)
      )
      
      pop_mean <- ifelse(!is.null(fit$Beta), fit$Beta$Estimate[1], 0)
      
      predictions <- gxe_df %>%
        dplyr::filter(ENV == e & GID %in% focal_gids) %>%
        dplyr::left_join(main_df, by = "GID") %>%
        dplyr::mutate(Predicted_YLD = pop_mean + Main_Effect + GxE_Effect)
      
      # 6. Calculate Accuracy
      eval_df <- env_data %>% dplyr::filter(GID %in% focal_gids) %>%
        dplyr::left_join(predictions, by = c("ENV", "GID")) %>%
        dplyr::filter(!is.na(Predicted_YLD) & !is.na(YLD))
      
      if(nrow(eval_df) > 3){
        r_val <- cor(eval_df$YLD, eval_df$Predicted_YLD, use = "complete.obs")
        cv_results <- rbind(cv_results, data.frame(Trial = tri, Target_Env = e, Fold = f, Pearson_r = r_val, Valid_N = nrow(eval_df)))
      }
      rm(K_ge, fit, He_sub, Kg_sub, pheno_combined, eval_df, predictions)
    }
  }
}
write.csv(cv_results, "Strict_CV00_Results.csv", row.names = FALSE)
write("CV00 Complete!", stdout())
```

### CV00 Final Predictions

```{r}
library(parallel)
library(dplyr)

# 1. Define the number of safe cores to use (9 trials = 9 cores)
safe_cores <- 8

write(paste0("Starting CV00 parallel processing using ", safe_cores, " cores..."), stderr())

# 2. Convert your loop into a function
process_trial_cv00 <- function(tri) {
  gc() 
  
  target_sub <- target_pheno %>% dplyr::filter(trial == tri)
  
  if(!tri %in% names(grmList)){
    return(NULL) 
  }
  
  gid.geno <- rownames(grmList[[tri]])
  geno_sub <- grmList[[tri]]
  
  target_sub <- target_sub %>% dplyr::filter(GID %in% gid.geno)
  if(nrow(target_sub) == 0){
    return(NULL)
  }
  
  # Identify the specific targets we need to completely blind the model to
  target_gids <- unique(target_sub$GID)
  target_envs <- unique(target_sub$ENV)
  
  # =====================================================================
  # THE STRICT CV00 FILTER
  # Start with all historical data that matches our GRM
  pheno_clean <- pheno %>% dplyr::filter(GID %in% gid.geno)
  
  # Wipe out ALL records of the target environments
  # Wipe out ALL records of the target genotypes
  pheno_clean <- pheno_clean %>%
    dplyr::filter(!ENV %in% target_envs) %>%
    dplyr::filter(!GID %in% target_gids)
  # =====================================================================
  
  # The RAM Saver Filter
  robust_historical_envs <- pheno_clean %>%
    dplyr::group_by(ENV) %>%
    dplyr::summarize(n = n(), .groups = "drop") %>%
    dplyr::filter(n >= 30) %>%  
    dplyr::pull(ENV)
  
  rescue_env <- "Neoga_2022" 
  
  target_envs <- unique(target_sub$ENV)
  
  # Force the rescue environment into the allowed list so it doesn't get dropped
  envs_to_keep <- unique(c(target_envs, robust_historical_envs, rescue_env))
  
  pheno_clean <- pheno_clean %>% dplyr::filter(ENV %in% envs_to_keep)
  
  # Target-specific filter for YT_Urb_25
  if(tri == "YT_Urb_25"){
    pheno_clean <- pheno_clean %>%
      dplyr::mutate(numeric_year = as.numeric(as.character(YEAR)))
    
    recent_years <- sort(unique(pheno_clean$numeric_year), decreasing = TRUE)[1:2]
    
    # Keep the last 2 years OR the specific rescue environment
    pheno_clean <- pheno_clean %>%
      dplyr::filter(numeric_year %in% recent_years | ENV == rescue_env) %>%
      dplyr::select(-numeric_year)
  }
  
  # Explicitly force target yields to NA to prevent any accidental leakage
  target_na <- target_sub %>% 
    dplyr::select(-trial) %>%
    dplyr::mutate(YLD = NA)
  
  pheno_combined <- dplyr::bind_rows(pheno_clean, target_na)
  
  missing_envs <- setdiff(target_envs, rownames(He))
  if(length(missing_envs) > 0){
    return(NULL)
  }
  
  valid_envs <- intersect(unique(pheno_combined$ENV), rownames(He))
  pheno_combined <- pheno_combined %>% dplyr::filter(ENV %in% valid_envs)
  all_gids <- unique(pheno_combined$GID) 
  
  # Build Matrices
  He_sub <- He[valid_envs, valid_envs]
  Kg_sub <- geno_sub[all_gids, all_gids]
  
  K_ge <- kronecker(He_sub, Kg_sub)
  ge_names <- paste(rep(rownames(He_sub), each = length(all_gids)),
                    rep(rownames(Kg_sub), times = length(rownames(He_sub))), sep = ":::")
  rownames(K_ge) <- colnames(K_ge) <- ge_names
  
  pheno_combined <- pheno_combined %>% 
    dplyr::mutate(GE_ID = factor(paste(ENV, GID, sep = ":::"), levels = rownames(K_ge)))
  
  # Run the Model 
  fit <- tryCatch({
    sommer::mmer(fixed = YLD ~ LOC + YEAR, 
                 random = ~ sommer::vsr(GID, Gu = Kg_sub) + sommer::vsr(GE_ID, Gu = K_ge), 
                 rcov = ~ units, 
                 data = pheno_combined,
                 verbose = FALSE) 
  }, error = function(e){
    return(NULL)
  })
  
  if(is.null(fit)) return(NULL)
  
  # Extract Main Genotype Effect
  blups_main <- fit$U$`u:GID`[[1]]
  if(is.null(blups_main)) blups_main <- fit$U[[1]][[1]]
  
  main_df <- data.frame(
    GID = gsub("^GID", "", names(blups_main)),
    Main_Effect = as.numeric(blups_main)
  )
  
  # Extract GxE Effect
  blups_gxe <- fit$U$`u:GE_ID`[[1]]
  if(is.null(blups_gxe)) blups_gxe <- fit$U[[2]][[1]]
  
  gxe_df <- data.frame(
    GE_ID = names(blups_gxe),
    GxE_Effect = as.numeric(blups_gxe)
  ) %>%
    dplyr::mutate(
      ENV = sub(":::.*", "", GE_ID),
      GID = sub(".*:::", "", GE_ID)
    )
  
  pop_mean <- mean(pheno_clean$YLD, na.rm = TRUE)
  
  # Combine predictions
  trial_predictions <- gxe_df %>%
    dplyr::left_join(main_df, by = "GID") %>%
    dplyr::mutate(Predicted_YLD = pop_mean + Main_Effect + GxE_Effect,
                  standard.error = NA) %>%
    dplyr::select(ENV, GID, Predicted_YLD, standard.error)
  
  # Filter back down to the target accessions
  trial_export <- target_sub %>%
    dplyr::left_join(trial_predictions, by = c("ENV", "GID"))
  
  return(trial_export)
}

# 3a. Separate the massive trial from the safe ones
safe_trials <- target_trials[target_trials != "YT_Urb_25"]

# 3b. Run the safe trials in parallel
write("Running standard trials in parallel...", stderr())
results_safe <- mclapply(safe_trials, process_trial_cv00, mc.cores = safe_cores)

# 3c. Run YT_Urb_25 completely by itself with MAXIMUM RAM
write("Running massive YT_Urb_25 trial sequentially (Max RAM)...", stderr())
result_urb <- process_trial_cv00("YT_Urb_25")

# 4. Bind everything together safely
# We wrap result_urb in a list() so bind_rows accepts it alongside the parallel list
all_final_predictions_cv00 <- dplyr::bind_rows(results_safe, list(result_urb))

write.csv(all_final_predictions_cv00, "CV00_Final_Yield_Predictions.csv", row.names = FALSE)
write("CV00 Processing complete!", stderr())
```

## CV00

Objective: predict unknown genotypes in unknown trials in cross-validation. For each trial in a given program, identify the genotypes and remove them from the training set (or remove a subset of it if small number of genotypes). Train the model with other genotypes tested in other environments.

```{r gblup_cv00}
res.cv00 <- expand.grid(method="GBLUP", scenario="CV00",
                        trial=target.pred,  corP=NA,wEnv.corP=NA,
                        nb.genos=NA, nb.envs=NA)
obs.pred.all <- data.frame()

for(tri in target.pred){
  write(paste0("Processing trial: ", tri,"\n"), stderr())
  
  ## subset GID with geno and pheno data
  gid.geno <- rownames(grmList[[tri]])
  gid <- intersect(gid.geno, pheno$GID)
  length(gid)
  if(length(gid) == 0){
    write(paste0("No genotypes with both geno and pheno data for trial ", tri), stdout())
    
  }else {
    ## subset GID with geno and pheno data
    gid.geno <- rownames(grmList[[tri]])
    gid <- intersect(gid.geno, pheno$GID)
    length(gid)
    ## subset pheno and geno from the program to predict
    geno_sub <- grmList[[tri]][gid,gid]
    pheno_sub <- pheno |> filter(GID %in% gid)
    df_pred_e <- data.frame()
    
    ## select environments with more than 30 obs
    envs <- pheno_sub %>%
      group_by(ENV) %>% summarize(n=n(),.groups="drop") %>%
      filter(n > 30) %>% pull(ENV)
    #pheno_sub <- pheno_sub %>% filter(ENV %in% envs)
    if(length(envs) == 0){
      write(paste0("No environment with more than 30 observations for trial ", tri), stdout())
      next
    } else {
      
      ## for loop over environments to predict each environment in turn
      for(e in envs){
        ## set yield data for the test environment to NA +
        ### set genotypes in the test environment as NA for all other environments
        gid.env <- pheno_sub$GID[pheno_sub$ENV %in% e]
        pheno_train <- pheno_sub |>
          mutate(YLD=ifelse(ENV %in% e, NA, YLD),
                 YLD=ifelse(GID %in% gid.env, NA, YLD)) %>%
          filter(GID %in% rownames(geno_sub)) %>%
          as.data.frame()
        #pheno_train <- as.data.frame(pheno_train[match(gid, pheno_train$GID),])
        #stopifnot(identical(rownames(geno_sub), pheno_train$GID))
        ## fit GBLUP model
        ### add fixed effects of location and year instead of env?
        fit <- rrBLUP::kin.blup(data=pheno_train,
                                geno="GID",pheno="YLD",
                                K=geno_sub,
                                fixed=c("ENV"),#"ENV", ## add env as fixed effect
                                n.core=nb.cores)
        ## for rrBLUP, predicted values averaged over fixed effects
        pred.all <- data.frame(ENV=e,
                               GID=names(fit$pred), pred=c(fit$pred))
        ## export predicted values for the test environment
        df_pred_e <- rbind(df_pred_e,filter(pred.all, ENV %in% e))
      }
      ## add observed values
      df_pred_e <- df_pred_e |> left_join(pheno_sub, by=c("GID","ENV")) %>% 
        mutate(target.trial=tri,
               scenario="CV00",
               GP.method="GBLUP")
      obs.pred.all <- rbind(obs.pred.all, df_pred_e)
      ## plot obs vs. pred
      plot(ggplot(df_pred_e, aes(x=YLD, y=pred)) +
             geom_point(alpha=0.6, size=1) +
             geom_smooth(method="lm", se=FALSE, color="black")+
             geom_rug(sides="bl")+
             #paletteer::scale_color_paletteer_d("ggthemes::Jewel_Bright") +
             ggpubr::stat_cor(method="pearson", size=3)+
             geom_abline(intercept=0, slope=1, color="red")+
             ggtitle(paste0("Predicted vs. observed values for trial ", tri),
                     subtitle=paste0("GBLUP genomic selection method"))+
             theme_bigstatsr(size.rel=0.8))
      ## calculate PA in each environment
      pa.within.env <- df_pred_e %>% group_by(ENV) %>%
        summarize(pa=cor(YLD, pred, use="pairwise.complete.obs"))
      ## calculate correlation across environments
      pa <- round(cor(df_pred_e$YLD, df_pred_e$pred, use="pairwise.complete.obs"),3)
      idx <- which(res.cv00$trial == tri)
      res.cv00$corP[idx] <- round(pa,3)
      res.cv00$wEnv.corP[idx] <- mean(pa.within.env$pa, na.rm=TRUE)
      res.cv00$nb.envs[idx] <- length(envs)
      ## add training/validation set size of the last fold
      res.cv00$nb.genos[idx] <- length(unique(pheno_train$GID[!is.na(pheno_train$YLD)]))
    }
  }
  
}
print_table(res.cv00)
write.table(res.cv00, here("output","GP_GBLUP_CV00_combined-GRM_common-pheno.tsv"),
            sep="\t", row.names=FALSE)

saveRDS(obs.pred.all, file=here("output","GP_GBLUP_CV00_combined-GRM_common-pheno_obs-pred.rds"))
```

# RKHS

Examples of BGLR models: <https://github.com/gdlc/BGLR-R/blob/master/inst/md/RKHS.md>

## CV0

```{r rkhs_cv0}
dir.create(here("output","temp"))
res.cv0 <- expand.grid(method="RKHS", scenario="CV0",
                       trial=target.pred,  corP=NA,
                       wEnv.corP=NA,nb.genos=NA, nb.envs=NA)
obs.pred.all <- data.frame()

for(tri in target.pred){
  write(paste0("Processing trial: ", tri,"\n"), stderr())
  
  ## subset GID with geno and pheno data
  gid.geno <- rownames(grmList[[tri]])
  gid <- intersect(gid.geno, pheno$GID)
  length(gid)
  if(length(gid) == 0){
    write(paste0("No genotypes with both geno and pheno data for trial ", tri), stdout())
    
  }else {
    ## subset pheno and geno from the program to predict
    geno_sub <- grmList[[tri]][gid,gid]
    pheno_sub <- pheno |> filter(GID %in% gid)
    ## select environments with more than 30 obs
    envs <- pheno_sub %>%
      group_by(ENV) %>% summarize(n=n(),.groups="drop") %>%
      filter(n > 30) %>% pull(ENV)
    #pheno_sub <- pheno_sub %>% filter(ENV %in% envs)
    if(length(envs) == 0){
      write(paste0("No environment with more than 30 observations for trial ", tri), stdout())
      next
    } else {
      df_pred_e <- data.frame()
      ## for loop over environments to predict each environment in turn
      for(e in envs){
        ## set yield data for the test environment to NA
        pheno_train <- pheno_sub |>
          mutate(YLD=ifelse(ENV %in% e, NA, YLD))
        pheno_train <- as.data.frame(pheno_train[match(gid, pheno_train$GID),])
        stopifnot(identical(rownames(geno_sub), pheno_train$GID))
        ## fit  BGLR RKHS
        Zg <- model.matrix(~GID, pheno_train)
        D<-as.matrix(dist(geno_sub,method="euclidean"))^2
        D<-D/mean(D)
        h<-1
        K<-exp(-h*D)
        Kg <- Zg %*% K %*% t(Zg)
        ## W: envt cov
        ## Zw = model.matrix(~ENV, pheno_train)
        ## Kw <- Zw %*% W %*% t(Zw)
        if(length(envs) < 3){
          fit <- BGLR::BGLR(y=pheno_train$YLD,
                            ETA=list(
                              #env=list(~factor(LOC)+factor(YEAR), data=pheno_train, model="BRR"),
                              g=list(K=Kg, model="RKHS")),
                            nIter=6000, burnIn=1500,
                            saveAt =here("output","temp",paste0("RKHS_loc-env_CV0_tri-",tri,"_env-",e)))
        } else {
          fit <- BGLR::BGLR(y=pheno_train$YLD,
                            ETA=list(
                              env=list(~factor(LOC)+factor(YEAR),
                                       data=pheno_train, model="BRR"),
                              #e=list(K=Kw, model="RKHS"),
                              g=list(K=Kg, model="RKHS")),
                            nIter=6000, burnIn=1500,
                            saveAt =here("output","temp",paste0("RKHS_loc-env_CV0_tri-",tri,"_env-",e)))
        }
        
        ## export predicted values
        #stopifnot(identical(names(fit$pred), pheno_train$GID))
        pred.all <- data.frame(ENV=pheno_train$ENV,
                               GID=pheno_train$GID, pred=c(fit$yHat))
        ## export predicted values for the test environment
        df_pred_e <- rbind(df_pred_e,filter(pred.all, ENV %in% e))
      }
      ## add observed values
      df_pred_e <- df_pred_e |> left_join(pheno_sub, by=c("GID","ENV")) %>%
        mutate(target.trial=tri,
               scenario="CV0",
               GP.method="RKHS")
      obs.pred.all <- rbind(obs.pred.all, df_pred_e)
      ## plot obs vs. pred
      plot(ggplot(df_pred_e, aes(x=YLD, y=pred, color=ENV)) +
             geom_point(alpha=0.6, size=1) +
             geom_smooth(method="lm", se=FALSE)+
             geom_rug(sides="bl")+
             #paletteer::scale_color_paletteer_d("ggthemes::Jewel_Bright") +
             ggpubr::stat_cor(method="pearson", size=3)+
             geom_abline(intercept=0, slope=1, color="grey", linetype="dashed")+
             ggtitle(paste0("Predicted vs. observed values for trial ", tri),
                     subtitle=paste0("GBLUP genomic selection method"))+
             theme_bigstatsr(size.rel=0.8))
      ## calculate PA in each environment
      pa.within.env <- df_pred_e %>% group_by(ENV) %>%
        summarize(pa=cor(YLD, pred, use="pairwise.complete.obs")) %>%
        as.data.frame()
      ## calculate correlation across environments
      pa <- round(cor(df_pred_e$YLD, df_pred_e$pred, use="pairwise.complete.obs"),3)
      idx <- which(res.cv0$trial == tri)
      res.cv0$corP[idx] <- round(pa,3)
      res.cv0$wEnv.corP[idx] <- round(mean(pa.within.env$pa, na.rm=TRUE),3)
      res.cv0$nb.envs[idx] <- length(envs)
      ## add training/validation set size of the last fold
      res.cv0$nb.genos[idx] <- length(unique(pheno_train$GID[!is.na(pheno_train$YLD)]))
    }
  }
}
print_table(res.cv0)
write.table(res.cv0, here("output","GP_RKHS_CV0_combined-GRM_common-pheno.tsv"),
            sep="\t", row.names=FALSE)
saveRDS(obs.pred.all, file=here("output","GP_RKHS_CV0_combined-GRM_common-pheno_obs-pred.rds"))
```

## CV00

```{r rkhs_cv00}
res.cv00 <- expand.grid(method="RKHS", scenario="CV00",
                        trial=target.pred,  corP=NA,wEnv.corP=NA,
                        nb.genos=NA, nb.envs=NA)
obs.pred.all <- data.frame()


for(tri in target.pred){
  write(paste0("Processing trial: ", tri,"\n"), stderr())
  
  ## subset GID with geno and pheno data
  gid.geno <- rownames(grmList[[tri]])
  gid <- intersect(gid.geno, pheno$GID)
  length(gid)
  if(length(gid) == 0){
    write(paste0("No genotypes with both geno and pheno data for trial ", tri), stderr())
    
  }else {
    ## subset GID with geno and pheno data
    gid.geno <- rownames(grmList[[tri]])
    gid <- intersect(gid.geno, pheno$GID)
    length(gid)
    ## subset pheno and geno from the program to predict
    geno_sub <- grmList[[tri]][gid,gid]
    pheno_sub <- pheno |> filter(GID %in% gid)
    df_pred_e <- data.frame()
    
    ## select environments with more than 30 obs
    envs <- pheno_sub %>% 
      group_by(ENV) %>% summarize(n=n(),.groups="drop") %>% 
      filter(n > 30) %>% pull(ENV)
    #pheno_sub <- pheno_sub %>% filter(ENV %in% envs)
    if(length(envs) == 0){
      write(paste0("No environment with more than 30 observations for trial ", tri), stdout())
      next
    } else {
      
      ## for loop over environments to predict each environment in turn
      for(e in envs){
        ## set yield data for the test environment to NA +
        ### set genotypes in the test environment as NA for all other environments
        gid.env <- pheno_sub$GID[pheno_sub$ENV %in% e]
        pheno_train <- pheno_sub |> 
          mutate(YLD=ifelse(ENV %in% e, NA, YLD),
                 YLD=ifelse(GID %in% gid.env, NA, YLD)) %>% 
          filter(GID %in% rownames(geno_sub)) %>% 
          as.data.frame()
        
        ## BGLR RKHS
        Zg <- model.matrix(~GID, pheno_train)
        D<-as.matrix(dist(geno_sub,method="euclidean"))^2
        D<-D/mean(D)
        h<-1
        K<-exp(-h*D)
        Kg <- Zg %*% K %*% t(Zg)
        
        fit <- BGLR::BGLR(y=pheno_train$YLD,
                          ETA=list(
                            env=list(~factor(LOC)+factor(YEAR), data=pheno_train, model="BRR"),
                            #env=list(~factor(LOC)+factor(YEAR), data=pheno_train, model="FIXED"),
                            g=list(K=Kg, model="RKHS")),
                          nIter=6000, burnIn=1500,
                          saveAt =here("output","temp",paste0("RKHS_loc-env_CV00_tri-",tri,"_final_")))
        
        ## export predicted values
        #stopifnot(identical(names(fit$pred), pheno_train$GID))
        pred.all <- data.frame(ENV=pheno_train$ENV, 
                               GID=pheno_train$GID, pred=c(fit$yHat))
        ## export predicted values for the test environment
        df_pred_e <- rbind(df_pred_e,filter(pred.all, ENV %in% e))
      }
      ## add observed values
      df_pred_e <- df_pred_e |> left_join(pheno_sub, by=c("GID","ENV")) %>% 
        mutate(target.trial=tri,
               scenario="CV00",
               GP.method="RKHS")
      obs.pred.all <- rbind(obs.pred.all, df_pred_e)
      ## plot obs. vs pred
      plot(ggplot(df_pred_e, aes(x=YLD, y=pred, color=ENV)) + 
             geom_point(alpha=0.6, size=1) +  
             geom_smooth(method="lm", se=FALSE, color="black")+
             geom_rug(sides="bl")+
             #paletteer::scale_color_paletteer_d("ggthemes::Jewel_Bright") + 
             ggpubr::stat_cor(method="pearson", size=3)+ 
             geom_abline(intercept=0, slope=1, color="grey",linetype="dashed")+
             ggtitle(paste0("Obs. vs. pred. values for trial ", tri), 
                     subtitle=paste0("RKHS genomic selection method"))+
             theme_bigstatsr(size.rel=0.8))
      ## calculate PA in each environment
      pa.within.env <- df_pred_e %>% group_by(ENV) %>% 
        summarize(pa=cor(YLD, pred, use="pairwise.complete.obs"))
      ## calculate correlation across environments
      pa <- round(cor(df_pred_e$YLD, df_pred_e$pred, use="pairwise.complete.obs"),3)
      idx <- which(res.cv00$trial == tri)
      res.cv00$corP[idx] <- round(pa,3)
      res.cv00$wEnv.corP[idx] <- mean(pa.within.env$pa, na.rm=TRUE)
      res.cv00$nb.envs[idx] <- length(envs)
      ## add training/validation set size of the last fold
      res.cv00$nb.genos[idx] <- length(unique(pheno_train$GID[!is.na(pheno_train$YLD)]))
    }
  }
  
}
print_table(res.cv00)
write.table(res.cv00, here("output","GP_RKHS_CV00_combined-GRM_common-pheno.tsv"),
            sep="\t", row.names=FALSE)
saveRDS(obs.pred.all, file=here("output","GP_RKHS_CV00_combined-GRM_common-pheno_obs-pred.rds"))

```

# Final predictions

## Submission requirements

Folder Structure:

The zipped folder submitted should contain one methods description text file and 9 sub-directories, each named with the trial name for which the sub-directory contains predictions. For example, if the sub-directory contains predictions for the trial "AYT_Timbuktu24", then the sub-directory should be named "AYT_Timbuktu24". Each sub-directory should contain six csv files. Continuing the example, the sub-directory should contain:

```         
CV0_Predictions.csv
CV0_Trials.csv
CV0_Accessions.csv
CV00_Predictions.csv
CV00_Trials.csv
CV00_Accessions.csv
```

These files should be formatted as follows.

CSV file formats:

```         
CV0_Predictions.csv and CV00_Predictions.csv: csv file with two columns, "germplasmName" and "prediction". All accessions in the focal trial that are genotyped should have a prediction.
CV0_Trials.csv and CV00_Trials.csv: a csv file with a single column, "studyName", containing the trial names of the trials used for training the prediction model.
CV0_Accessions.csv and CV00_Accessions.csv: a csv file with a single column, "germplasmName", containing the accession names used for training the prediction model.
```

```{r}
# 1. Load the target accessions
target_data <- read_csv(here("data","acc_for_prediction.csv"))

target_pheno <- target_data %>%
  dplyr::rename(GID = acc_name,
                LOC = location,
                YEAR = year) %>%  
  dplyr::mutate(YLD = NA) %>%
  dplyr::select(trial, ENV, GID, LOC, YEAR, YLD)

all_final_predictions <- data.frame()
target_trials <- unique(target_pheno$trial)
```


## CV0

```{r CV0_final, include=FALSE, echo=FALSE}
CV0 <- data.frame()
for(tri in target_trials){
  write(paste0("Processing trial: ", tri,"\n"), stderr())
  env.pred <- target_pheno |> filter(trial %in% tri) |> 
    dplyr::select(trial, ENV, LOC, YEAR) |> distinct()
  pred.gid <- target_pheno |> filter(trial %in% tri) |> pull(GID)
  # p2d <- here("submission", tri)
  # dir.create(p2d)
  ## subset GID with geno and pheno data
  gid.geno <- rownames(grmList[[tri]])
  gid <- intersect(gid.geno, pheno$GID)
  length(gid)
  
  ## subset pheno and geno from the program to predict
  
  pheno_sub <- pheno |> filter(GID %in% gid)
  
  ## set yield data for the test environment to NA
  pheno_train <- pheno_sub |>
    bind_rows(target_pheno %>% filter(trial %in% tri)) %>% 
    as.data.frame() %>% filter(GID %in%  gid.geno)
  gid <- intersect(pheno_train$GID, gid.geno)
  geno_sub <- geno_sub <- grmList[[tri]][gid,gid]
  
  ## output accessions and trials used for training
  trialsTrain <- data.frame(studyName=unique(pheno_sub$study_name))
  germplasmTrain <- data.frame(germplasmName=unique(pheno_sub$GID))
  
  p2f.tri <- here("submission",tri,"CV0_Trials.csv")
  write.table(trialsTrain, file=p2f.tri, sep=",", col.names=TRUE, row.names=F)
  
  p2f.geno <- here("submission", tri, "CV0_Accessions.csv")
  write.table(germplasmTrain, file=p2f.geno, sep=",", col.names=TRUE, row.names=F)

  # ## fit GBLUP model
  # ### add fixed effects of location and year instead of env?
  fit <- rrBLUP::kin.blup(data=pheno_train, geno="GID",pheno="YLD",
                          K=geno_sub,
                          fixed=c("LOC","YEAR"),#"ENV", ## add env as fixed effect
                          n.core=nb.cores)
  pred.all <- data.frame(GID=names(fit$pred), pred=c(fit$pred))
  pred.sub <- pred.all |> rename(germplasmName=GID,
                                 prediction=pred) %>%
    filter(germplasmName %in% pred.gid)

  p2f <- here("submission",tri, "CV0_Predictions.csv")
  write.table(pred.sub, p2f, sep=",", col.names=TRUE, row.names=FALSE)
  CV0 <- rbind(CV0, pred.all)
}
saveRDS(CV0, file=here("output","GP_RKHS_CV0_final.rds"))
```




## CV00

```{r message=FALSE, warning=FALSE, include=FALSE}
CV00 <- data.frame()
for(tri in target_trials){
  write(paste0("Processing trial: ", tri,"\n"), stderr())
  env.pred <- target_pheno |> filter(trial %in% tri) |> 
    dplyr::select(trial, ENV, LOC, YEAR) |> distinct()
  gid.pred <- target_pheno |> filter(trial %in% tri) |> pull(GID)
  ## subset GID with geno and pheno data
  gid.geno <- rownames(grmList[[tri]])
  gid <- intersect(gid.geno, pheno$GID)
  
  ## set yield data for the test environment to NA
  pheno_sub <- pheno |> filter(GID %in% gid.geno, !GID %in% gid.pred)
  pheno_train <- pheno_sub |>
    bind_rows(target_pheno %>% filter(trial %in% tri)) %>% 
    as.data.frame() %>% filter(GID %in%  gid.geno)
  gid <- intersect(pheno_train$GID, gid.geno)
  geno_sub <- geno_sub <- grmList[[tri]][gid,gid]
  
  pheno_train$YLD[pheno_train$ENV %in% env.pred$ENV] <- NA
  pheno_train$YLD[pheno_train$GID %in% gid.pred] <- NA
  ## output accessions and trials used for training
  trialsTrain <- data.frame(studyName=unique(pheno_sub$study_name))
  germplasmTrain <- data.frame(germplasmName=unique(pheno_sub$GID))
  
  p2f.tri <- here("submission",tri,"CV00_Trials.csv")
  write.table(trialsTrain, file=p2f.tri, sep=",", col.names=TRUE, row.names=F)
  
  p2f.geno <- here("submission", tri, "CV00_Accessions.csv")
  write.table(germplasmTrain, file=p2f.geno, sep=",", col.names=TRUE, row.names=F)
  
  ## subset pheno and geno from the program to predict
  geno_sub <- grmList[[tri]][gid,gid]
  
  ## BGLR RKHS
  Zg <- model.matrix(~GID, pheno_train)
  D<-as.matrix(dist(geno_sub,method="euclidean"))^2
  D<-D/mean(D)
  h<-1
  K<-exp(-h*D)
  Kg <- Zg %*% K %*% t(Zg)

  fit <- BGLR::BGLR(y=pheno_train$YLD,
                    ETA=list(
                      env=list(~factor(LOC)+factor(YEAR), data=pheno_train, model="BRR"),
                      #env=list(~factor(LOC)+factor(YEAR), data=pheno_train, model="FIXED"),
                      g=list(K=Kg, model="RKHS")),
                    nIter=10000, burnIn=2000,
                    saveAt =here("output","temp",paste0("RKHS_loc-env_CV00_tri-",tri,"_final_")))

  ## export predicted values
  pred.all <- data.frame(ENV=pheno_train$ENV,
                         GID=pheno_train$GID, pred=c(fit$yHat)) |>
    filter(ENV %in% env.pred$ENV)
  pred.sub <- pred.all |> select(-ENV) |> rename(germplasmName=GID,
                                                 prediction=pred)
  # p2d <- here("submission", tri)
  # dir.create(p2d)
  p2f <- here("submission",tri, "CV00_Predictions.csv")
  write.table(pred.sub, p2f, sep=",", col.names=TRUE, row.names=FALSE)
  CV00 <- rbind(CV00, pred.all)
}
saveRDS(CV00, file=here("output","GP_RKHS_CV00_final.rds"))

```

## Verifications


Verify that we have everything
```{r}
for(tri in target.pred){
  pheno2pred <- target_pheno |> filter(trial %in% tri)
  pred.cv0 <- read.table(file=here("submission",tri, "CV0_Predictions.csv"),
                         sep=",",header=TRUE)
  sd <- setdiff(pheno2pred$GID, pred.cv0$germplasmName)
  if(length(sd)){
    grm <- grmList[[tri]]
    if(any(sd %in% rownames(grm))) {
      print(paste0("missing geno CV0 for trial ", tri,": ",setdiff(rownames(grm), sd)))
    }
  }
  pred.cv00 <- read.table(file=here("submission",tri, "CV00_Predictions.csv"),
                          sep=",",header=TRUE)
  sd <- setdiff(pheno2pred$GID, pred.cv00$germplasmName)
  if(length(sd)){
    grm <- grmList[[tri]]
    if(any(sd %in% rownames(grm))) {
      print(paste0("missing geno CV00 for trial ", tri,": ",setdiff(rownames(grm), sd)))
    }
  }
  
}
```

# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
