---
title: "Predictathon"
author: "Prediction5 Team"
date: "`r Sys.Date()`"
output: html_document
---

# Libraries

```{r}
suppressPackageStartupMessages(library(BrAPI)) # T3 connection)
suppressPackageStartupMessages(library(purrr)) # handle lists
suppressPackageStartupMessages(library(tidyverse)) # data handling
suppressPackageStartupMessages(library(here)) # root directory
suppressPackageStartupMessages(library(apercu)) # look at lists
suppressPackageStartupMessages(library(progress)) # progress bar
suppressPackageStartupMessages(library(data.table)) # fread read table

```

# Set up root directory

TODO: need to move this script to analysis folder

```{r current dir}
here::i_am("analysis/phenotypic_data.Rmd")
source(here("code","useful_functions.R"))
```

# WheatCAP Predictathon: Phenotypic Data

## Accessing T3 Breedbase

```{r}
rm(list=ls())

wheat <- createBrAPIConnection("wheat.triticeaetoolbox.org", is_breedbase=TRUE)
wheat <- getBrAPIConnection("T3/Wheat")

#wheat$get("/germplasm")
focal_trials <- c("AWY1_DVPWA_2024", "TCAP_2025_MANKS", "25_Big6_SVREC_SVREC",
                  "OHRWW_2025_SPO", "CornellMaster_2025_McGowan", "24Crk_AY2-3",
                  "2025_AYT_Aurora","YT_Urb_25","STP1_2025_MCG")
```

## Accessing Trial Metadata

```{r}


check_val <- function(x) {
  if (is.null(x) || length(x) == 0) return(NA)
  return(x)
}
results_list <- list()
for(trial_name in focal_trials){
  resp <- wheat$get("/studies", query=list(studyName=trial_name))
  if (length(resp$data) == 0) {
    message(paste("Skipping: No data found for", trial_name))
    next
  }
  trial_metadata <- resp$data[[1]]
  row_data <- data.frame(
    trial_name    = trial_name,
    program       = check_val(trial_metadata$trialName),
    trial_id      = check_val(trial_metadata$studyDbId),
    location      = check_val(trial_metadata$locationName),
    year         = check_val(unlist(trial_metadata$seasons)),
    planting_date = check_val(trial_metadata$startDate),
    harvest_date  = check_val(trial_metadata$endDate),
    design        = check_val(trial_metadata$experimentalDesign$PUI),
    stringsAsFactors = FALSE
  )
  results_list[[trial_name]] <- row_data
}
focal_metadata <- do.call(rbind, results_list)
```

## CV00: Gather Program Phenotypic Data

### Grabbing all Trials for our Programs of Interest

```{r}
programs <- focal_metadata$program
all_trials_list <- list()
for (prog in programs) {
  resp <- wheat$get("/studies", query = list(programName=prog, pages="all", pageSize=1000))
  if (length(resp$data) == 0) next
  ## create a data frame summarizing trial id, name, location, year, program
  prog_trials <- lapply(resp$data, function(x) {
    data.frame(studyDbId = x$studyDbId, 
               name = x$studyName, 
               location = x$locationName,
               year=paste(unlist(x$seasons), collapse=", "), 
               program = prog)
  })
  prog_trials <- do.call(rbind, prog_trials)
  ## if there are too many trials, select those from the same year and location as focal trial
  if(nrow(prog_trials) > 20){
    focal_years <- focal_metadata %>% filter(program == prog) |> pull(year)
    focal_locations <- focal_metadata %>% filter(program == prog) %>% pull(location) %>% unique()
    prog_trials <- prog_trials %>% 
      filter(year %in% focal_years | location %in% focal_locations)
  }
  
  all_trials_list[[prog]] <- prog_trials
}
## look first five elements
ap(all_trials_list)
lapply(all_trials_list, nrow)
all_trials <- bind_rows(all_trials_list, .id="focal_trial")
all_trials$isFocal <- ifelse(all_trials$name %in% focal_trials, TRUE, FALSE)
table(all_trials$isFocal)
all(focal_trials %in% all_trials$name)

## export trial names to pull genotypic data from it
p2f <- here("data","allTrials_metadata.csv") # path to file
write.table(all_trials, p2f, sep=",", row.names=F, quote=F)

all_trials <- all_trials %>% filter(!name %in% focal_metadata$trial_name)
trial_ids <- all_trials$id
vars_rm <- "season"
```

### Yield Data for All Trials

```{r}

# 
# all_observations <- data.frame()
# 
# pb <- txtProgressBar(min=1, max=length(trial_ids), style = 3)
# ## this trial id is long: 10703
# ## this trial id times out: 10710, 10876
# trial_ids <- trial_ids[!trial_ids %in% c("10703","10710","10876")]
# 
# for(i in seq_along(trial_ids)){
#   setTxtProgressBar(pb, i)
#   #write(paste0("Processing trial ", i, " of ", length(trial_ids)), stdout())
#   search <- try(wheat$search("/observations", body=list(studyDbIds = c(trial_ids[i]))))
#   results <- search$combined_data
#   results <- results %>%
#     map(function(item){
#       item[vars_rm] <- NULL
#       return(item)
#     }) %>% 
#     bind_rows
#   if(nrow(results) >0){
#     results <- results %>% filter(observationVariableName == "Grain yield - kg/ha|CO_321:0001218")
#     all_observations <- bind_rows(all_observations, results)
#   }
# }
# close(pb)
# 
# ## export all trials yield
# p2f <- here("data","phenotypic","allTrials_yield_raw.csv") # path to file
# write.table(all_observations, p2f, sep=",", row.names=F, quote=F)
```

Modifications compared to previous version:

-   Print the iteration number to track progress and see if some trials are causing problems
-   Test if trial data has yield variable
-   Replace `rbind` with `bind_rows` for handling when columns names differ

## CV0: Pull Phenotype Data for Focal Accessions

Alternatives to load yield data more efficiently (?)

1.  Use the \$get

```         
resp <- wheat$get("/observations",
query=list(germplasmDbId=focal_ids[1],
observationVariableDbIds = list("84527"),
studyDbId=all_trials$id[1]), 

page="all", pageSize=500)
```

2.  Use the search

```         
search <- try(wheat$search("/observations", body=list(
germplasmDbIds = c(focal_ids), 
observationVariableDbIds = list("84527"),
studyDbIds=all_trials$id,
pageSize=10000
)))
```

```{r}
p2f <- here("data","phenotypic","allAccessions_yield_raw.csv") 
if(file.exists(p2f)){
  results <- fread(p2f, header=T, sep=",")
} else {
  
  wheat <- getBrAPIConnection("T3/WheatCAP")
  focal_acc <- wheat$wizard("accessions", list(trials=c(focal_trials)), verbose=T)
  focal_acc <- as.data.frame(cbind(focal_acc$data$ids, focal_acc$data$names))
  colnames(focal_acc)[1] = "acc_id"
  colnames(focal_acc)[2] = "acc_name"
  focal_ids <- focal_acc$acc_id
  wheat <- getBrAPIConnection("T3/Wheat")
  vars_rm <- "season"
  
  search <- try(wheat$search("/observations", body=list(
    #germplasmDbIds = c(focal_ids), 
    observationVariableDbIds = list(84527),
    studyDbIds=all_trials$studyDbId,
    # page="all",
    pageSize=100000
  )))
  
  
  results <- search$combined_data
  results <- results %>%
    map(function(item){
      item[vars_rm] <- NULL
      return(item)
    }) %>% 
    bind_rows
  ## export all accessions yield
  write.table(results, p2f, sep=",", row.names=F, quote=F)
}

```

Add the trial metadata in that file if not already present

```{r}
colN <- colnames(results)
if(!"location" %in% colN){
  results <- results |> 
    mutate(studyDbId = as.character(studyDbId)) |> 
    left_join(all_trials, by=c("studyDbId"="studyDbId"))
  ## export all accessions yield with trial metadata
  write.table(results, p2f, sep=",", row.names=F, quote=TRUE, col.names=TRUE)
}
#rm(all_trials_list, focal_acc, focal_metadata)
```

# Combine and filter phenotypic data from J-L Jannink

For each target trial, select phenotypic data for the genotyped lines and add trial metadata (location, start/end dates) to it. Then combine phenotypic data from all trials into one file.

```{r}
for(tri in focal_trials){
  #tri=focal_trials[1]
  all.files <- list.files(here("data","phenotypic",tri))
  ## for now, select all phenotype files
  sel.files <- all.files[grepl("phenotypes", all.files)]
  p2fs <- paste0(here("data","phenotypic",tri),"/",sel.files)
  dat.tri <- map_dfr(p2fs, fread)
  ap(dat.tri)
  
  ## add the trial location & year
  p2f.met <- paste0(here("data","phenotypic",tri),"/",grep("metadata", all.files, value=T))
  trial.info <- fread(p2f.met, header=T, sep=",")
  
  ## load combined geno for this trial
  p2f <- here("data","genotypic", tri,"clean",
              paste0(tri,"_combined_GRM.rds"))
  geno.tri <- readRDS(p2f)
  gid.geno.tri <- rownames(geno.tri)
  length(gid.geno.tri)
  length(intersect(unique(dat.tri$germplasm_name), gid.geno.tri))
  
  ## select phenotypic data from the genotyped inds
  dat.tri.sel <- dat.tri |> 
    filter(germplasm_name %in% gid.geno.tri) |> 
    left_join(trial.info |> 
                select(study_db_id,location_name, study_name,program_name),
              by=c("study_db_id")) |> 
    rename(year=season)
  
  write(paste0("Trial ", tri, ": ", dim(dat.tri.sel)[1], " phenotypic records for ",
               length(unique(dat.tri.sel$germplasm_name)), " genotyped lines"), stdout())
  p2f.save <- here("data","phenotypic",tri,
              paste0(tri,"_pheno_genoSelected.csv"))
  write.table(dat.tri.sel, p2f.save, sep=",", row.names=F, quote=T, col.names=TRUE)
}

## combine phenotypic data into one list
pheno.all <- map_dfr(focal_trials, function(tri){
  p2f <- here("data","phenotypic",tri,
              paste0(tri,"_pheno_genoSelected.csv"))
  dat <- fread(p2f, header=T, sep=",") |> 
    mutate(trial.target=tri)
  return(dat)
})
write.table(pheno.all, here("data","phenotypic","allTrials_pheno_genoSelected.csv"), 
            sep=",", row.names=F, quote=T, col.names=TRUE)
```
For 24Crk_AY2-3, we have less lines than with the initial phenotypic data file, for all other trials we have more genotypes.

