---
title: "Pull genotype data from T3"
author: "Prediction5 Team"
output: html_document
date: '`r format(Sys.time(), "%d %B, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

# Libraries

```{r load libraries}
suppressPackageStartupMessages(library(here)) # root directory
suppressPackageStartupMessages(library(data.table)) # load data
suppressPackageStartupMessages(library(apercu)) # ap
suppressPackageStartupMessages(library(dplyr)) # formatting
suppressPackageStartupMessages(library(purrr)) # handle lists
#devtools::install_github("TriticeaeToolbox/BrAPI.R")
#pak::pak("jeanlucj/T3BrapiHelpers")
suppressPackageStartupMessages(library(BrAPI)) # T3 connection
suppressPackageStartupMessages(library(T3BrapiHelpers)) # T3 helper
suppressPackageStartupMessages(library(DT)) # print table
suppressPackageStartupMessages(library(ggplot2)) # plot
suppressPackageStartupMessages(library(vcfR)) # load vcf
suppressPackageStartupMessages(library(progress)) # progress bar
suppressPackageStartupMessages(library(janitor)) # make clean names

```

# Set up root directory

```{r current dir}
here::i_am("analysis/pullGenoAssociatedTrials.Rmd")
source(here("code","useful_functions.R"))
```

```{r}
trials_predictathon <- c("AWY1_DVPWA_2024","TCAP_2025_MANKS","25_Big6_SVREC_SVREC",
                         "OHRWW_2025_SPO","CornellMaster_2025_McGowan","24Crk_AY2-3",
                         "2025_AYT_Aurora","YT_Urb_25","STP1_2025_MCG")
```

# Create T3 connection

What's the difference between getBrAPIConnection and createBrAPIConnection?

```{r}
#wheat <- getBrAPIConnection("T3/Wheat")
db <- getBrAPIConnection("T3/Wheat")
wheat <- createBrAPIConnection("wheat.triticeaetoolbox.org", is_breedbase=TRUE)
#trials_df <- T3BrapiHelpers::getAllTrialMetaData(wheat,"Wheat")
```

# Get genotyping platform used for each focal trial

```{r}
p2ds <- here("data","genotypic",trials_predictathon)
list.files(p2ds)
## load manifest files for each focal trial
manifests <- map_dfr(trials_predictathon, \(x){
  fread(here("data","genotypic",x,"manifest.csv"))
})
manifests
protid = unique(manifests$protocol_id)
```

# Get list of associated trials

```{r}
p2f <- here("data","allTrials_metadata.csv") # path to file
all_trials_init <- fread(p2f) |> 
  select(-V1) 
all_trials_focal <- all_trials_init |> 
  filter(isFocal) 
focal_prg <- unique(all_trials_focal$program)

all_trials_init <- all_trials_init |> filter(!isFocal)
head(all_trials_init)
```

# Pull genotyping platform for each trial

Trial by trial

```{r pull_geno_platform, include=TRUE}

p2f <- here("data","allTrials_metadata_genotyping_platform.csv") # path to file

# if(file.exists(p2f)){
#   all_trials <- fread(p2f) 
# } else {
  all_trials <- all_trials_init |> 
    bind_cols(nb_genos = NA, prop_geno = NA, best_protocol = NA, 
              best_protocol_id = NA, best_project = NA, best_project_id = NA,
              vcf_file_name = NA, inManifest = FALSE)
  #tri=all_trials$focal_trial[165]
  pb <- txtProgressBar(min=1, max=nrow(all_trials), style = 3)
  for(i in 1:nrow(all_trials)){
    setTxtProgressBar(pb, i)
    tri <- all_trials$focal_trial[i]
    ## get the list of genotypes in this trial
    #genos <- getGermplasmFromTrialVec(study_id_vec=unique(wsu$focal_trial), db)
    genos <- getGermplasmFromTrialVec(study_id_vec=tri, db)
    
    ## add the total number of lines
    all_trials$nb_genos[i] <- nrow(genos)
    ## which genotyping platform is the best?
    resp <- try(db$filter_geno_projects(genos$germplasmDbId))
    if(length(resp) == 4 && resp$status$category == "Success"){
      top_match <- resp$data$top_matches
      ## TODO if there are multiple options, take the one in common with the focal trial
      #all_trials$best_project[i] <- top_match[1]
      
      ## from the project name, get the id
      projId <- resp$content$results$counts$ranked_genotyping_projects[[1]]
      #all_trials$best_project_id[i] <- projId
      
      ## get the protocol name the same way
      req <- db$filter_geno_protocols(accessions = genos$germplasmDbId, verbose = FALSE)
      protId <- req$content$results$counts$ranked_genotyping_protocols[[1]]
      
      info <- wheat$vcf_archived_list(genotyping_protocol_id=protId, verbose = FALSE)
      if(nrow(info) > 1){
        info <- info |> filter(project_id == projId, protocol_id == protId) |> 
          head(1) |> as.data.frame()
      }
      if(nrow(info) == 0){
        print(paste0("No VCF found for trial ", tri, " with protocol id ", protId, " and project id ", projId))
        next
      }
      all_trials$best_protocol[i] <- info$protocol_name
      all_trials$best_protocol_id[i] <- info$protocol_id
      all_trials$best_project[i] <- info$project_name
      all_trials$best_project_id[i] <- info$project_id
      all_trials$vcf_file_name[i] <- info$file_name
      if(info$file_name %in% manifests$file_name){
        all_trials$inManifest[i] <- TRUE
      }
      
      ## get the proportion of lines genotyped 
      nb_genos <- sum(resp$data$matches[,top_match[1]], na.rm=T)
      all_trials$prop_geno[i] <- 100*round(nb_genos / nrow(genos),1)
      # ## add the list of genotypes in the trial
      # all_trials$genos_list[i] <- list(genos$germplasmDbId)
    } else {
      all_trials$best_protocol[i] <- NA
    }
  }
  close(pb)
  ## Export the table with genotyping platform info
  write.table(all_trials, p2f, sep=",", row.names=F, quote=F)
#}
```

# Explore output

Number of trials with genotyped lines for each program

```{r}
all_trials_geno <- all_trials |> 
  filter(!is.na(prop_geno) & prop_geno > 1 & nb_genos > 5) 
dim(all_trials_geno) ## 61 x 14
table(all_trials_geno$program)
```

Kansas doesn't have any trial with genotyped lines.

Check the genotyping platform diversity

```{r}
table(all_trials_geno$best_protocol)
intersect(manifests$protocol_name, unique(all_trials_geno$best_protocol))
```

# Download all vcf files

```{r}

#for(prog in focal_prg[!focal_prg %in% "USDA-ARS Kansas"]){
for(prog in focal_prg){
  prog_trials <- all_trials_geno |> 
    filter(program == prog) |> 
    select(best_protocol_id, best_project_id, vcf_file_name, inManifest) |> 
    distinct()
  if(nrow(prog_trials) == 0) {
    write(paste0("No genotyped trial found for program ", prog), stdout())
  } else {
    
    
    write(paste0("Downloading ", nrow(prog_trials), " VCF files for program ", prog), stdout())
    tri <- unique(all_trials_focal |> filter(program == prog) |> pull(studyDbId))
    p2d <- here("data","genotypic",tri,"nonFocal")
    if(!dir.exists(p2d)) dir.create(p2d, recursive = TRUE)
    
    better_prog_name <- make_clean_names(prog,case="snake")
    for(i in 1:nrow(prog_trials)){
      if(!prog_trials$inManifest[i]){## Don't download if same file as focal trial
        wheat$vcf_archived(
          output =paste0(p2d,"/",better_prog_name,"_",i,".vcf.gz"),
          genotyping_protocol_id = prog_trials$best_protocol_id[i],
          genotyping_project_id =  prog_trials$best_project_id[i],
          file_name = prog_trials$vcf_file_name[i],
          verbose = FALSE
        )
      } else{
        write(paste0("VCF file ", prog_trials$vcf_file_name[i], " for trial ", 
                     prog, " is already downloaded"), stdout())
        
      }
    }
  }
}
```
