---
title: "Environmental_data_import"
author: "Prediction3 Team"
date: "2025-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This script takes the output .csv document ([genotyped_phenotyped_trials.csv](genotyped_phenotyped_trials.csv)) from [T3_trial_import_and_filtering.Rmd](T3_trial_import_and_filtering.Rmd) and uses it as an input into EnvRtype package functions to pull historical weather data for each trial. The output of this script are .csv files that give raw weather data for ~30 parameters across the growing season for each trial selected for our project.

## Load Packages

```{r Load Packages, message=FALSE}
library(devtools) #install.packages("devtools")
install_github("TriticeaeToolbox/BrAPI.R")
library(BrAPI)
library(here) #install.packages("here")
library(dplyr) #install.packages("dplyr")
library(janitor) #install.packages("janitor")
library(tidyverse) #install.packages("tidyverse")
library(readr) #install.packages("readr")
library(reshape2) #install.packages("reshape2")
devtools::install_github("allogamous/EnvRtype")
library(EnvRtype)
library(kableExtra) #install.packages("kableExtra")
library(maps) #install.packages("maps")
```

## Set BrAPI Server

```{r Create BrAPI Connection, message=FALSE}
conn <- createBrAPIConnection("wheat.triticeaetoolbox.org", is_breedbase=TRUE)
```

## Load Data

This is the output data frame from our trial filtering in [T3_trial_import_and_filtering.Rmd](T3_trial_import_and_filtering.Rmd)

```{r Load Data, message=FALSE}
selected_trials <- readr::read_csv(here("data","genotyped_phenotyped_trials.csv"))
```

## Pull T3 Trial Location Data

For each trial on T3, uploaders assign a location. We can query T3 to get metadata about each of these locations, including longitude, latitude, and elevation.

```{r T3 Location data pull, message=FALSE,warning=FALSE}
trial_locations <- unique(selected_trials$location_name)

loc_search <- conn$post("search/locations",body=list(
  locationNames = trial_locations
))
loc_response <- conn$get(paste0("search/locations/",loc_search$content$result$searchResultsDbId),pageSize=10000)
loc_result <- loc_response$content$result$data

loc_df <- bind_rows(lapply(loc_result, function(x) {
  coords <- unlist(x$coordinates$geometry$coordinates)
  
  data.frame(
    locationDbId = x$locationDbId %||% NA,
    locationName = x$locationName %||% NA,
    abbreviation = x$abbreviation %||% NA,
    latitude = if (!is.null(coords) && length(coords) >= 2) coords[2] else NA,
    longitude = if (!is.null(coords) && length(coords) >= 1) coords[1] else NA,
    elevation = if (!is.null(coords) && length(coords) >= 3) coords[3] else NA,
    stringsAsFactors = FALSE
  )
}))

selected_trials <- dplyr::left_join(selected_trials,loc_df, by=c("location_name"="locationName"))

rm(loc_search)
rm(loc_response)

kable(loc_df, format = "html") |> 
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |> 
  scroll_box(width = "800px", height = "500px")
```

```{r map of trial locations}
us_contig <- map_data("state") %>% filter(!region %in% c("alaska","hawaii"))

ggplot() +
  geom_polygon(data = us_contig,
               aes(x = long, y = lat, group = group),
               fill = "white", color = "darkgray") +
  coord_fixed(ratio = 1.3,
              xlim = c(-130, -65),
              ylim = c(22.5,   50))+
  geom_point(data = loc_df,
             aes(x = longitude, y = latitude),
             fill="darkgreen",color = "black", size = 5, shape = 25)+
  geom_text(data = loc_df,
             aes(x = longitude, y = latitude, label=abbreviation),
             color = "black", size = 2.5, vjust=4)+
  labs(
    title="Locations of selected T3 trials")+
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(hjust = 0.5)
  )
```

## Pull NASA-Generated Weather Data

### Function

This function uses the [**EnvRtype** package](https://doi.org10.1093/g3journal/jkab040) to query the [NASA POWER](https://power.larc.nasa.gov/) (Prediction of Worldwide Energy Resources) database system to estimate a set of historical radiation, temperature, and precipitation parameters at a set of locations that we provide from the latitudinal and longitudinal cooridinates we just pulled from T3. 

```{r get_trial_weather_data function}
get_trial_weather_data <- function(trial_df, col_locationName, col_startDate, col_endDate, col_latitude, col_longitude){
  
  weather_output <- list()
  
  for (i in 1:nrow(trial_df)){
    current_trial <- trial_df[i,]
    trial_weather <- get_weather(env.id=current_trial[[col_locationName]],
                                 lat=current_trial[[col_latitude]],
                                 lon=current_trial[[col_longitude]],
                                 start.day =current_trial[[col_startDate]] ,
                                 end.day =current_trial[[col_endDate]] 
                                 )
    weather_output[[i]] <- trial_weather
  }
  bind_rows(weather_output)
}
```


Now we can run this function for our list of study_db_ids.

*This takes a few minutes to run*
```{r get weather data, message=FALSE, cache=TRUE}
selected_trials$study_db_id <- as.factor(selected_trials$study_db_id)
trial_weather_df <- get_trial_weather_data(selected_trials,"study_db_id","start_date","end_date","latitude","longitude")

readr::write_csv(trial_weather_df, here::here("data","EnvRtype_output.csv"))

kable(summary(dplyr::select(trial_weather_df,7:32)), format="html") |> 
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |> 
  scroll_box(width = "800px", height = "500px")
```

### Process Data

We can use the EnvRtype::processWTH() function to estimate some weather hyperparameters from our dataset.

```{r process weather data}
trial_weather_processed <- trial_weather_df |> processWTH()

trial_weather_processed |> left_join(selected_trials,by=c("env"="study_db_id")) |> 
  readr::write_csv(here::here("data","processed_weather_data.csv"))

kable(summary(dplyr::select(trial_weather_processed,7:42)), format="html") |> 
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |> 
  scroll_box(width = "800px", height = "500px")
```

## Soil data

We can also pull soil data right from EnvRtype for most environments

*this will take a few minutes to run*

```{r}
vars <- c("bdod","cec","cfvo","clay","nitrogen","ocd","soc","phh2o","sand","silt")
location_soil_data <- get_soil(env.id <- loc_df$locationName, lat = loc_df$latitude, lon=loc_df$longitude, variables.names = vars)
location_soil_data <- location_soil_data |> clean_names() |> select(environmental_unit,ends_with("mean"))
write_csv(location_soil_data, here::here("data","location_soil_data.csv"))

kable(location_soil_data, format="html") |> 
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |> 
  scroll_box(width = "800px", height = "500px")
```


## Pull Data for test trials

```{r test trial location data pull}
test_trials <- readr::read_csv(here("data","test_trial_metadata.csv")) |> 
  clean_names()

test_trial_locations <- unique(test_trials$location_name)

test_loc_search <- conn$post("search/locations",body=list(
  locationNames = test_trial_locations
))
test_loc_response <- conn$get(paste0("search/locations/",test_loc_search$content$result$searchResultsDbId),pageSize=10000)
test_loc_result <- test_loc_response$content$result$data

test_loc_df <- bind_rows(lapply(test_loc_result, function(x) {
  coords <- unlist(x$coordinates$geometry$coordinates)
  
  data.frame(
    locationDbId = x$locationDbId %||% NA,
    locationName = x$locationName %||% NA,
    abbreviation = x$abbreviation %||% NA,
    latitude = if (!is.null(coords) && length(coords) >= 2) coords[2] else NA,
    longitude = if (!is.null(coords) && length(coords) >= 1) coords[1] else NA,
    elevation = if (!is.null(coords) && length(coords) >= 3) coords[3] else NA,
    stringsAsFactors = FALSE
  )
}))

test_trials <- dplyr::left_join(test_trials,test_loc_df, by=c("location_name"="locationName"))

rm(test_loc_search)
rm(test_loc_response)

kable(test_loc_df, format = "html") |> 
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |> 
  scroll_box(width = "800px", height = "500px")
```

```{r test trial weather data pull}
test_trials$study_db_id <- as.factor(test_trials$study_db_id)

test_trials[test_trials$study_db_id==10674, "end_date"] = as.Date("2024-08-26") #harvest date was missing for this trial, so I imputed the harvest date from the trial in this location in 2022

test_trial_weather_df <- get_trial_weather_data(test_trials,"study_db_id","start_date","end_date","latitude","longitude")

readr::write_csv(test_trial_weather_df, here::here("data","testset_EnvRtype_output.csv"))

kable(summary(dplyr::select(test_trial_weather_df,7:32)), format="html") |> 
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |> 
  scroll_box(width = "800px", height = "500px")
```
```{r process test set weather data}
test_trial_weather_processed <- test_trial_weather_df |> processWTH()

test_trial_weather_processed |> left_join(test_trials,by=c("env"="study_db_id")) |> 
  readr::write_csv(here::here("data","testset_processed_weather_data.csv"))

kable(summary(dplyr::select(test_trial_weather_processed,7:42)), format="html") |> 
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |> 
  scroll_box(width = "800px", height = "500px")
```

```{r test set soil data pull}
vars <- c("bdod","cec","cfvo","clay","nitrogen","ocd","soc","phh2o","sand","silt")
test_location_soil_data <- get_soil(env.id <- test_loc_df$locationName, lat = test_loc_df$latitude, lon=test_loc_df$longitude, variables.names = vars)
test_location_soil_data <- test_location_soil_data |> clean_names() |> select(environmental_unit,ends_with("mean"))
write_csv(test_location_soil_data, here::here("data","test_location_soil_data.csv"))

kable(test_location_soil_data, format="html") |> 
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |> 
  scroll_box(width = "800px", height = "500px")
```

