---
title: "Phenotypic_data_import"
author: "Prediction3 Team"
date: "2025-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This script takes the output .csv document ([genotyped_phenotyped_trials.csv](genotyped_phenotyped_trials.csv)) from [T3_trial_import_and_filtering.Rmd](T3_trial_import_and_filtering.Rmd) and uses the selected study ids to query T3 and pull all phenotypic data for each trial. Additionally, to remove drone-derived traits from the phenotyping data, an input .csv document ([drone_traits_lookup.csv](drone_traits_lookup.csv)), is used that contains data pulled from the [T3 WheatCAP Drone Trait Lookup](https://wheatcap.triticeaetoolbox.org/guides/drone-traits) The output of this script is a csv file containing all phenotypic observations for each entry in our selected trials.

## Load Packages

```{r Load Packages, message=FALSE}
library(devtools) #install.packages("devtools")
install_github("TriticeaeToolbox/BrAPI.R")
library(BrAPI) 
library(purrr) #install.packages("purrr")
library(tidyverse) #install.packages("tidyverse")
library(readr) #install.packages("readr")
library(here) #install.packages("here")
```

## Set BrAPI Server

```{r Create BrAPI Connection, message=FALSE}
conn <- createBrAPIConnection("wheat.triticeaetoolbox.org", is_breedbase = TRUE)
```

## Load Data

This is the output data frame from our trial filtering in [T3_trial_import_and_filtering.Rmd](T3_trial_import_and_filtering.Rmd)

```{r Load Data, message=FALSE}
selected_trials <- readr::read_csv(here("data","genotyped_phenotyped_trials.csv"))
selected_trial_ids <- selected_trials$study_db_id
```

## Pull T3 Trial Phenotype Data

### Function

This function queries T3 and pulls all "observations" (phenotypic records) for each of our selected studies

```{r function to query T3 for phenotypic observations}
get_obs_for_studies <- function(conn, study_ids) {
  `%||%` <- function(x, y) if (!is.null(x)) x else y
  all_data <- list()
  
  for (id in study_ids) {
    message("Getting observations for studyDbId: ", id)
    
    obs_search <- tryCatch({
      conn$post("search/observations", body = list(studyDbIds = id))
    }, error = function(e) {
      warning("Failed for studyDbId ", id)
      return(NULL)
    })
    
    if (is.null(obs_search)) next
    
    if (!is.null(obs_search$content$result$data)) {
      obs_result <- obs_search$content$result$data
    } else {
      obs_search_id <- obs_search$content$result$searchResultsDbId
      page <- 0
      all_obs <- list()
      
      repeat {
        obs_response <- conn$get(
          paste0("search/observations/", obs_search_id),
          page = page,
          pageSize = 1000
        )
        
        this_page <- obs_response$content$result$data
        if (length(this_page) == 0) break
        
        all_obs <- c(all_obs, this_page)
        page <- page + 1
      }
      
      obs_result <- all_obs
    }
    
    if (length(obs_result) > 0) {
      pheno_df <- bind_rows(lapply(obs_result, function(x) {
        data.frame(
          studyDbId = x$studyDbId %||% NA,
          season = x$season$year %||% NA,
          observationUnitName = x$observationUnitName %||% NA,
          germplasmDbId = x$germplasmDbId %||% NA,
          germplasmName = x$germplasmName %||% NA,
          observationVariableName = x$observationVariableName %||% NA,
          observationUnitDbId = x$observationUnitDbId %||% NA,
          value = x$value %||% NA,
          stringsAsFactors = FALSE
        )
      }))
      all_data[[id]] <- pheno_df
    }
  }
  bind_rows(all_data)
}
```

### Pull Data

We can run this function for our selected study ids. After pulling the data, we pivot the table to have each trait set as a column and filter out columns associated with UAV-derived traits. 

*this takes a few minutes to run*

```{r pull and filter phenotype data,cache=TRUE}
all_obs_data <- get_obs_for_studies(conn,selected_trial_ids)

all_obs_data_pivot <- all_obs_data |> 
  pivot_wider(names_from = observationVariableName, values_from = value) |> 
  dplyr::select(-contains("COMP:")) |>  #REMOVES DRONE TRAITS
  janitor::clean_names()

selected_trials$study_db_id <- as.character(selected_trials$study_db_id)

all_obs_data_pivot <- all_obs_data_pivot |> 
  left_join(dplyr::select(selected_trials,study_db_id,study_name), by = "study_db_id")

readr::write_csv(all_obs_data_pivot,here("data","genotyped_trial_phenotypes.csv"))
```

